\section*{Reproducibility}
\label{sec:repro}

\if\anon1
All training, evaluation, and analysis code is released anonymously at
\url{https://anonymous.4open.science/r/matrix-thinking-chapter2}.
\else
All training, evaluation, and analysis code is released at
\url{https://github.com/saml212/learned-representations} under
\texttt{matrix-thinking/chapter2/}.
\fi

\paragraph{Code, by experiment.} Task D: \texttt{task\_d.py} (grammar
generator), \texttt{model\_v4.py} (encoder + pinned unbind),
\texttt{rank\_utils.py} (eigh-stable effective/stable rank),
\texttt{run\_task\_d.py} (train/eval/smoke), \texttt{run\_overnight.py}
(perpetual, resumable 8$\times$H100 orchestrator). Task E: \texttt{task\_e.py},
\texttt{run\_task\_e.py}, \texttt{run\_task\_e\_sweep.py} (reuses Task D's
encoder and rank utilities verbatim), \texttt{eigen\_utils.py}
(eigenstructure-fidelity metric, C8), \texttt{analyze\_zdump.py} (numpy +
stdlib only, no GPU required; regenerates every number in
\S\ref{sec:taske-subspace}--\ref{sec:taske-kwall} from archived $Z$-dumps).
Stage 0: \texttt{run\_stage0.py}, \texttt{run\_stage0\_sweep.py}.

\paragraph{Pre-registrations.} \texttt{TASK\_D\_PREREGISTRATION.md}
(hypothesis, proof, controls, decision criteria, logged before any
training); \texttt{NEXT\_EXPERIMENT\_DESIGN.md} (Task E design and its
$H_{\mathrm{extra}}$ residue-collision addendum); \texttt{STAGE0\_DESIGN.md}
(the exactness-frontier diagnostic, formal and diagnostic pass criteria).
Every pre-registered decision criterion and every deviation from it (e.g.\
the $\tau{=}0.99\to0.9$ metric re-registration in \S\ref{sec:taskd-m3},
logged before the sweep; Stage 0's Wave B skip in favor of the 100K probe,
\S\ref{sec:frontier-pattern}) is documented inline in these files with a
timestamp, not silently applied.

\paragraph{Audit trail.} Design and code passed an adversarial gauntlet
before any GPU run for each experiment: a task-shortcut attack (which
killed an earlier crossover design and produced Task D; which caught the
injectivity-tolerance and general-permutation bugs that would have silently
broken Task E's held-out-hop claim before any compute ran), a novelty
check, three rounds of code audit, and two rounds of runner audit
(\texttt{gauntlet/}), plus a runtime smoke gate (degenerate-spectrum
backward, blank-out leak test, resolution pre-flight).

\paragraph{Hardware and compute.} Brev 8$\times$H100 80GB cluster, torch
2.12/cu13. Task D: $\approx$56 GPU-h for the 991-run snapshot reported in
\S\ref{sec:taskd}, plus a subsequent 1{,}234-run mega-replication. Task E:
the 40K-step calibration round (33 runs), the 80K-step $K$-wall round (6
runs, 14.5 GPU-h), and the 120K-step $K{=}16$ completion wave (3 runs).
Stage 0: 27.66 GPU-h core (Wave $-1$/0/A, an extended-budget arm, and the
100K probe, against a pre-registered 60 GPU-h budget), plus an 11.36 GPU-h
post-closure exactness-frontier follow-on ($d{=}64$ and $d{=}48$ points).
Cumulative project grant burn at time of writing: $\approx$320 of a 1{,}600
GPU-h grant. The concurrent DeltaNet result in \S\ref{sec:discussion} is
$\approx$10 GPU-h so far, Waves $-1$/0 only, explicitly partial.

\paragraph{Data.} Raw per-run JSON results (dense training-step
checkpoints, not just final values -- every trajectory-based claim in this
paper, e.g.\ ``flat, not climbing'' vs.\ ``still climbing at cutoff,'' is
verified from the full checkpoint array, not a summary rollup) are archived
under \texttt{experiment-runs/}, in five dated subdirectories: \texttt{2026-07-02\_task\_e\_zdump},
\texttt{2026-07-02\_task\_e\_80k\_kwall}, \texttt{2026-07-02\_task\_e\_120k\_k16},
\texttt{2026-07-02\_stage0\_waves}, and \texttt{2026-07-02\_deltanet\_waves}.
All numbers in
this paper trace to a specific archived JSON, run tag, and seed; none are
recomputed or estimated for this write-up.
