\section{Experimental setup}
\label{sec:setup}

\subsection{Task: episode-restricted associative recall}

Each training and evaluation sequence is an \emph{episode}: $K = 32$ bind
clauses followed by a query. A bind clause is a 7-token statement linking a
key entity to a value entity, both drawn without replacement from a pool of
real GPT-2 token identifiers \citep{radford2019language}, so the binding
span is $T_{\mathrm{bind}} = K \times 7 = 224$
tokens. % <!-- evidence: C10 -->
The query names one key; the model must produce the bound value at the
answer position. Keys and values are re-drawn every episode, so the map is
in-context and cannot be memorized in slow weights. The primary task is
single-hop; a multi-hop variant appears in Section~\ref{sec:scope}.

\textbf{Metric.} We score \emph{episode-restricted recall accuracy}: the
argmax over the model's own LM-head logits restricted to the episode's $K$
bound values, read at the answer position of the model's own continuation.
Chance is $1/K = 0.03125$, and an arm counts as demonstrating recall in a
cell only above the pre-registered demonstration bar of $3/K = 0.09375$;
readings at or below the bar are scored no-recall, and differences between
two no-recall arms are treated as chance-level noise ordering, never as
separation. One caveat travels with every accuracy in this paper: under
argmax decoding, an associative state of rank one can already support on
the order of $d$ associations \citep{nichani2024understanding}, so recall
accuracy here never establishes rank, continuous recovery, or storage
capacity. Section~\ref{sec:mechanism} reports the continuous-readout
diagnostics separately.

\subsection{Three arms, matched budget}

\textbf{Matrix fast-weight model (the contender).} A two-block
decoder-style LM (residual width 256, one head per mixer) whose mixer holds
a $64 \times 64$ matrix state $S_t$ per block, updated by a delta rule
\citep{schlag2021linear, yang2024parallelizing}:
$S_t = S_{t-1}(I - \beta_t k_t k_t^{\top}) + \beta_t v_t k_t^{\top}$, with
per-token read $o_t = S_t q_t$. Its total recurrent state is
$2 \times 64 \times 64$ fp32 values, or 32{,}768 bytes, constant in context
length. % <!-- evidence: C10 -->

\textbf{Flat-vector recurrence (the additive control).} Identical
embedding table, output head, and feed-forward blocks; only the mixer
changes, to a gated elementwise vector recurrence
$s_t = s_{t-1} \odot g_t + v_t$ with read $o_t = s_t \odot q_t$, at matched
total parameter count via mixer width and $d_{\mathrm{state}} = 64$ pinned
equal to the contender's. This arm removes the multiplicative
outer-product update while preserving everything else, rather than
reshaping the same operator, since any $d^2$-dimensional vector can encode
a $d \times d$ matrix and structure matters only if the operations use it.
This control differs from the contender on three axes, not one: mixer
recurrence, raw state capacity (512 bytes versus 32{,}768,
% <!-- evidence: C10 -->
reported for completeness), and write conditioning; its update has no
dependence on the key $k_t$ at all, so it cannot perform content-addressed
binding under any training outcome, unlike vector-native binding schemes
that use a key-dependent operation such as circular convolution
\citep{plate1995holographic} or tensor-product binding
\citep{smolensky1990tensor}. Section~\ref{sec:recall} states which
reading of the resulting comparison the capacity and write-conditioning
gaps license.

\textbf{Transformer.} A two-layer GPT-2-style causal decoder with the same
tokenizer and width, evaluated in two roles: uncapped (its full KV cache),
and KV-capped with attention-sink-plus-FIFO eviction
\citep{xiao2023efficient} at a total cache budget pinned to
$M \times 32{,}768$ bytes across all layers and heads, swept over
$M \in \{1, 2, 4, 8, 16, 32\}$. % <!-- evidence: C10 -->

Parameter counts are \nparamscontender{} (contender), \nparamsablation{}
(flat-vector, $-0.007\%$), and \nparamstransformer{} (transformer,
$+2.8\%$). % <!-- evidence: C9 -->
All arms train from scratch for 20{,}000 steps at learning rate
$3 \times 10^{-4}$ with the same optimizer, batch schedule, auxiliary
losses, and episode stream; sampling seeds are paired across arms, so the
three arms of a given seed index see identical
episodes. % <!-- evidence: C9 -->

\subsection{Analysis protocol, frozen before the runs}

Margins, tiers, and analysis rules were registered in the project's
experiment registry and frozen before the multi-seed runs. The primary
comparison is contender versus flat-vector at the matched final
checkpoint: a win requires the contender to clear the demonstration bar
and the paired-seed 95\% $t$-interval on the accuracy difference (three
seeds, $\mathrm{df}=2$) to exclude a 0.30 margin. Two clauses matter for
reading this paper. First, the registry pins a \emph{degenerate-baseline}
rule: if the uncapped transformer itself fails the demonstration bar at
the primary cell, the capped-budget walk is definitionally degenerate,
every cap is trivially cleared, and the registered outcome is ``baseline
non-competitive at matched params/tokens,'' with no memory-multiplier tier
certified. Second, a joint-no-recall rule scores any cell where both
compared arms sit at the bar as a tie, never as a separation. Evaluation
uses a pinned held-out query set of 4{,}096 queries over 128 episodes per
cell, % <!-- evidence: C3 -->
identical across arms and seeds. Three seeds is the registered count for
the primary comparison; the protocol includes a seed-extension trigger
for any cell whose paired gap approaches the 0.30 margin, and none did
(Table~\ref{tab:main}), % <!-- evidence: C2 -->
so the count was not extended for this comparison. The multi-hop
variant's later nine-seed table (Section~\ref{sec:scope}) reflects a
different, seed-variance-driven extension under the same protocol, not a
change in standard. The protocol treats the primary
contender-versus-additive-control comparison as the single
pre-registered decision rule; the remaining reported intervals in this
paper (the per-$M$ capped-budget gaps, the pooled multi-hop interval,
per-seed bar comparisons) are descriptive and diagnostic rather than
additional hypothesis tests, so no multiple-comparison correction is
applied to them.
