\section{Recall separation at matched budget}
\label{sec:recall}

\begin{table}[t]
\centering
\small
\tableone
\caption{Episode-restricted recall accuracy on the primary cell ($K=32$,
$K/d=0.5$, binding span 224 tokens), per seed and mean, at the matched
final checkpoint (20{,}000 steps, identical data and optimizer; chance
$1/32 = 0.031$, demonstration bar $3/32 = 0.094$). The matrix fast-weight
model clears the bar in every seed; neither the parameter-matched
flat-vector recurrence nor the parameter-matched uncapped transformer ever
does. Both paired-seed 95\% $t$-intervals ($\mathrm{df}=2$) exclude the
pre-registered 0.30 margin. Accuracy is argmax-decoded and
episode-restricted, so it carries the capacity caveat of
Section~\ref{sec:setup}.}
\label{tab:main}
\end{table}

Table~\ref{tab:main} reports the primary comparison. The contender reads
0.9995, 1.0000, and 0.9990 across the three seeds, % <!-- evidence: C1 -->
at least $10\times$ the demonstration bar in every seed; the flat-vector
arm reads 0.0322 to 0.0369 and the transformer 0.0271 to 0.0293, all
within noise of chance and below the bar in every
seed. % <!-- evidence: C1 -->
The paired difference against the flat-vector arm is 0.9656 with a 95\%
interval of $(0.9582,\, 0.9729)$; against the transformer, 0.9712 with
$(0.9685,\, 0.9738)$. % <!-- evidence: C2 -->
Both intervals exclude the frozen 0.30 margin with more than 0.65 to spare
at the interval floor, and no seed straddle fires the registered
seed-extension trigger. % <!-- evidence: C2 -->

The readings this table supports are narrower than the raw gap suggests.
Under the shared training budget, the matrix fast-weight update reaches
ceiling while the additive control's training loss barely moves past its
first-500-step value for the remaining 97.5\% of the run
(Appendix~\ref{app:traincurve}); learning rate, warmup, and
auxiliary-loss weighting were shared across three architecturally
distinct mixers, not independently tuned per arm, so this paper does not
claim the additive update is \emph{architecturally incapable} of the
task, only that it does not reach it under this budget. A second reading
holds but is narrower than ``matrix beats vector'': the additive
control also holds $64\times$ less raw state and a write with no
dependence on the query key (Section~\ref{sec:setup}), so the comparison
shows \emph{a matrix state with a multiplicative, key-conditioned
write outperforms a smaller, non-key-conditioned vector state under a
shared training budget}; capacity and write-conditioning are not
disentangled by this ablation, and no resized or key-conditioned vector
control was run. The transformer column carries its registered phrasing,
\emph{non-competitive at matched params/tokens}: at this scale and budget
it does not learn the task at all, which is itself the axis-2 datum,
never extrapolated to differently-scaled or differently-trained attention
models. What the table does not license is any claim that a transformer
\emph{cannot} perform this task, or that outer-product structure
specifically, rather than capacity and key-conditioning together, is
what the additive control's failure isolates. Attention solves
associative recall well in other regimes \citep{arora2023zoology,
nichani2024understanding}; under this task distribution, tokenizer,
parameter count, and step count, the attention arm did not get there
while the fast-weight arm reached ceiling in every seed.
