\section{Introduction}
\label{sec:intro}

% Full argument flow: NARRATIVE.md \S3 "\S1 Introduction"

\begin{itemize}
  \item \textbf{Open on the tension.} Descriptive rank-measurement papers
    (Nazari \& Rusch, arXiv:2602.04852; Sun et al., arXiv:2602.02195, both
    Feb 2026) show fast-weight state rank is bounded by sequence length in
    pretrained LLMs, but never show SGD is \emph{forced} to use that rank,
    nor that the resulting operator is \emph{exact}. We close both gaps in
    one architecture family.
  \item \textbf{Contribution 1 (synthetic, CONFIRMED tier).} A provable
    $\mathrm{rank}(S_T) \geq K$ necessity result for exact-continuous
    recovery in a production fast-weight layer (DeltaNet,
    \texttt{chunk\_delta\_rule}), causally confirmed via a razor-cliff
    eval-truncation staircase: $k{=}31{\to}32$ jumps
    $0.9681{\to}1.0000$ at $h{=}1$, $0.3412{\to}1.0000$ at $h{=}21$
    ($d{=}64, K{=}32$). No inflation: entity-subspace rank $=$ whole-matrix
    rank $=K$ to four decimal places.
  \item \textbf{Contribution 2 (real data, CONFIRMED, qualified).} The SAME
    causal necessity result transfers to real GPT-2-tokenized language at
    the identical $(d,K)$ operating point (RD-1) --- \textbf{always stated
    with the ``graded, not razor'' qualifier}: ceiling reached exactly at
    $k{=}K$ at every $K\in\{8,16,24,32\}$ tested, but the transition window
    widens from 2--3 ranks ($K{=}8$) to 12--13 ranks ($K{=}32$), and the
    ceiling itself falls with $K$ (1.000 $\to$ 0.795).
  \item \textbf{Contribution 3 (the new phenomenon).} What the synthetic
    result cannot see: per-binding exactness, not capacity, is what
    degrades with $K$ on real data, compounding geometrically with
    composition depth ($\sim\varepsilon^h$).
  \item \textbf{Contribution 4 (mechanism).} A causal attribution to
    write-time key geometry (ruling out value geometry, NCE-crowding,
    kernel precision) via a controlled embedding-source intervention, plus
    an existence proof (surgical orthonormal-key pin, ``i-strong'') that
    the exact solution is architecturally reachable: $\mathrm{rec@0.9} =
    1.00/1.00/1.00$ at $h{=}1/2/3$, $K{=}32$, vs.\ learned baseline
    $0.78/0.26/0.05$.
  \item \textbf{Contribution 5 (the fix, demonstrated not just diagnosed).}
    Differentiable per-episode Newton-Schulz key orthogonalization,
    launch-gated on a pre-registered drift simulator. Clears the $K{=}16$
    minimum-publishable bar (3/3 admissible seeds, $h{=}4$: $0.42$--$0.47
    \to 0.95$--$1.00$ vs.\ bar $\geq 0.8$); improves $K{=}32$
    $\sim$45$\times$ ($0.009 \to 0.39$--$0.50$, 3/3 fully admissible on
    the \texttt{n\_iter=20} escalation, zero fallback steps), narrowly
    missing the $\geq 0.5$ headline bar (mean $0.44$, 1/3 seeds at the
    line) --- with the miss itself attributed to a second,
    pre-registered, pre-measured mechanism (cross-episode key drift
    $0.9037$ in the registered HIGH band, outcome F). The causal chain
    closes with no unexplained residual.
  \item \textbf{Contribution 6 (the fix's own hypothesized follow-on
    mechanism, tested and REJECTED).} The residual outcome-F gap
    (cross-episode key drift) motivates a natural follow-on: a learned,
    per-entity cross-episode key anchor. It produces a real,
    independently-replicated $K{=}32$ $h{=}4$ behavioral gain
    ($0.4105 \to 0.61$--$0.71$, 9 seed-runs across 3 seed sets, 2
    architecture variants) --- but its hypothesized entity-alignment
    mechanism is REJECTED (Outcome C): a purpose-built per-entity
    engagement instrument reads $<$14--41\% engagement at every measured
    leg, confirmed across two independent mechanism-tier waves at
    increasing evidentiary rigor, immune to a specific rescue attempt that
    was itself independently attacked and rejected.
  \item \textbf{Contribution 7 (the actual explanation, found by
    pre-registered ablation).} A frozen, never-trained anchor table
    (random or frame-potential init) at the same blend weight matches or
    exceeds the learned table's own gain ($0.67$--$0.76$ vs.\
    $0.61$--$0.71$) --- constancy in the blend arithmetic, not
    entity-specific learned content, is the active ingredient. This is
    CONFIRMED-BY-ABLATION: a falsifiable prediction (``a frozen table
    should match the gain if constancy alone is the channel'') was stated
    before the confirming data existed and held on fresh data.
  \item \textbf{Contribution 8 (the capacity boundary, located, dissolved,
    and its leading confound exonerated).} Extending the identical fix to
    $K{=}48$ ($K/d{=}0.75$) shows both of its own named explanatory
    channels measurably active yet the fix still misses its bar 0/3
    seeds --- the capacity curve is a cliff
    ($\sim$1.00 $\to$ $\sim$0.65 $\to$ $\sim$0.02), not a smooth decline.
    A dedicated localization wave fits this transition's midpoint to
    $x_0{=}0.5455$ (95\% CI $[0.5385,0.5513]$) at $d_{\mathrm{state}}{=}64$;
    the identical $K/d$ window is then shown to DISSOLVE, not merely
    shift, at $d_{\mathrm{state}}{=}128$ ($h4{=}1.0$, 12/12 cells); the
    leading single-variable account of that dissolution (anchor-table
    coherence) is directly tested by injecting controlled, frozen
    coherence past $d{=}64$'s own realized band and is EXONERATED
    ($h4{=}1.0$, 10/10 cells). A further wave tests the remaining
    raw-state-capacity candidate directly at $d_{\mathrm{state}}{=}80/96$
    and finds load tolerance grows SUPER-LINEARLY, not at a fixed $K/d$
    ratio ($x_0(80){=}0.6779$, CI $[0.6683,0.6867]$, excluding the
    ratio-invariant band; a wide-grid $d{=}96$ attempt, after an
    independently-audited detour resolving a tolerance-miscalibration
    false alarm in its own admission gate, finds NO cliff through
    $K/d{=}0.9375$ but a non-monotonic curve that does not localize,
    100\% bootstrap-degenerate) --- the precise functional form remains
    open, stated as such rather than resolved.
  \item \textbf{Contribution 9 (process, stated once, lightly).} Every
    wave above was pre-registered, independently attacked, and --- for the
    fix --- gated on a predicted outcome from a committed simulator before
    any GPU spend. Offered as a template for trusting a negative result
    (Wave F) enough to publish it, not as a self-congratulatory framing
    device. Cite the two concrete bug catches (target-index gather bug,
    \S\ref{sec:phenomenon}; cross-episode key drift, \S\ref{sec:the-fix})
    as evidence, not the process in the abstract.
  \item \textbf{What this paper is NOT.} Not a claim about production
    Gated-DeltaNet at scale (H=1, single layer, sub-1M params here); not a
    claim that natural-language reasoning \emph{requires} rank $K$ (Wave 2
    is descriptive+interventional only); not a claim that the h=21
    depth-decay mechanism is fixed by this paper's intervention (it is
    not, stated explicitly in \S\ref{sec:results}).
\end{itemize}
