\section{Setup and Theory}
\label{sec:setup}

% Full grounding: DELTANET_CAUSAL_RANK_DESIGN.md \S1, \S2, \S4.1;
% NARRATIVE.md \S3 "\S3 Setup and Theory"

\subsection{Architecture}

Every experiment in this paper uses the same vehicle: a vanilla,
single-head ($H{=}1$), single-layer DeltaNet built on \texttt{fla-org}'s
production \texttt{chunk\_delta\_rule} kernel, with the standard delta
rule $S_t = S_{t-1}(I - \beta_t k_t k_t^\top) + \beta_t v_t k_t^\top$
governing how the $d\times d$ fast-weight state $S_t$ is written at every
step. $H{=}1$ and single-layer are a deliberate gate-simplicity choice,
not a claim that the phenomenon is $H{=}1$-specific: they close the
multi-head/multi-layer pigeonhole escape (rank distributed invisibly
across heads or layers) cleanly, at the cost of leaving the general case
untested here. Where that cost is already partially paid, we say so
rather than assume it: ReserveMH shows every head independently recruits
full rank $K{=}32$ at $H\in\{2,4\}$ with no cross-head distribution
(6/6 cells), lifting the $H{=}1$ qualifier on the synthetic necessity
task, but this generality check has not yet been re-run on the real-data
exactness frontier that is this paper's main subject --- carried forward
as a limitation in Section~\ref{sec:discussion}.

\subsection{The Provable Bound}

Every readout in this paper is pinned to the same linear-unbind form,
$\mathrm{pred} = S_T \cdot k_j$, scored by cosine similarity against the
target value at threshold $\tau{=}0.9$, and \emph{never} by argmax over a
codebook. This choice is not incidental: \citet{nichani2025factual} show
that under discrete argmax decoding a rank-1 state can recover
$\approx d$ associations, which would silently defeat any rank-necessity
claim built on this architecture. Pinning the readout to exact,
continuous recovery is what makes the following bound non-vacuous.

\begin{proposition}
\label{prop:rank-necessity}
Let $S_T\in\R^{d\times d}$ be a fast-weight state and
$\{(k_j,v_j)\}_{j=1}^{K}$ a set of $K$ bindings with $\{v_j\}$ linearly
independent. If a linear-unbind readout exactly recovers every binding,
i.e.\ $S_T k_j = c_j v_j$ for some $c_j\neq0$, then $\mathrm{rank}(S_T)
\geq K$.
\end{proposition}

\begin{proof}[Proof sketch]
Cosine scoring is invariant to a positive per-item scalar, so exact
recovery at $\tau{=}1$ only requires $S_T k_j$ to lie on the ray through
$v_j$ --- a learned $\beta_j\neq0$ suffices, it need not equal $1$. The
image of $S_T$ therefore contains $\{c_jv_j\}_{j=1}^K$, and since
$\{v_j\}$ are linearly independent and every $c_j\neq0$, so are
$\{c_jv_j\}$. A $K$-dimensional linearly independent set lying in
$\mathrm{im}(S_T)$ forces $\dim\,\mathrm{im}(S_T) = \mathrm{rank}(S_T)
\geq K$.
\end{proof}

Two escapes are closed by construction rather than left to the reader's
trust. First, position-decomposition: the state is a single $d\times d$
matrix, not a stack of $P$ independently addressable slots, so the bound
cannot be satisfied cheaply by parking each binding in its own rank-1
slot (a $P{=}1$ hard bottleneck, verified elsewhere in this project's
own gradient-based blank-out tests). Second, the architecture-specific
reincarnations of the same escape are closed by the setup itself:
multi-head distribution is closed by the $H{=}1$ primary gate above, and
multi-layer distribution is closed by construction --- the readout reads
only the final layer's state.

\subsection{Method: the Pre-Registration Discipline}

Every claim in this paper was pre-registered --- hypothesis and success
bar stated before data collection --- attacked by an independent
reviewer round, and, for the constructive fix
(Section~\ref{sec:the-fix}), gated on a predicted outcome from a
committed simulator before any Wave~1 GPU spend. We do not narrate this
discipline as a virtue in the abstract; we cite the specific findings it
produced wherever they did real work. A concrete instance: a mini-audit
caught a target-index bug in the real-data grammar (\texttt{forward()}
gathered $\pi^{h+1}(a)$ instead of the queried $\pi^h(a)$) that made the
binding task unsatisfiable as originally specified. The resulting
$\mathrm{rec@0.9}{=}1.000$ was not a positive result but a
value-collapse artifact --- entity-subspace rank $\approx1.0$, value
salvage $\approx0.000$, 0/10 seeds premise-valid --- caught by the
premise-gate instruments before it was ever written up as a finding, and
fixed with a pre-registered anti-collapse loss $L_{\mathrm{train}} =
L_{\cos} + \lambda_{\mathrm{nce}}\cdot L_{\mathrm{nce}}$ (in-episode
InfoNCE, $\lambda_{\mathrm{nce}}{=}1.0$, $T{=}0.1$, fixed not tuned),
chosen because collapse is a zero-gradient stationary point under the
NCE term. A second concrete instance, in the fix's own gating diagnostic,
appears in Section~\ref{sec:the-fix}.

Two distinct cells recur throughout this paper as ``the'' primary $K$,
and we disambiguate them once here rather than let context carry the
distinction. $K{=}16$ is the causal-necessity program's primary cell,
chosen because the unconstrained baseline's own $K{=}32$ rerun already
showed a real, non-budget sub-exactness plateau at
$\mathrm{rec@0.9}{=}0.78$--$0.80$ (Section~\ref{sec:phenomenon}).
$K{=}32$ is, precisely because it is that anomaly --- sitting at
$K{=}d_{\mathrm{state}}/2$ --- the exactness-mechanism and fix program's
headline cell. Both choices are principled and were pre-registered before
the waves that used them.
