\section{Soft Fixes Fail: The Attractor Is Robust}
\label{sec:soft-fixes}

% Full grounding: DELTANET_RD_EXACTNESS_DESIGN.md \S15 (18/18 cells);
% EXPERIMENT_LOG.md "Exactness Wave F (soft arms)" 2026-07-04.

\subsection{Two pre-registered interventions, one registered shot each}
\begin{itemize}
  \item F-geo-1: $L_{\mathrm{orth}}$ penalty, $\lambda \in \{0.1, 1.0\}$.
  \item F-geo-2: population-level ZCA whitening.
  \item Both ran to completion with NO further tuning (anti-Goodhart cap
    honored) --- state explicitly so it doesn't read as ``we could have
    tried harder.''
\end{itemize}

\subsection{Result: one attractor, three approach angles (Fig.~6)}
\begin{itemize}
  \item ALL land in the SAME narrow basin: gram dev $2.51$--$2.57$ at
    $K{=}32$ vs.\ baseline $2.71$--$2.77$ (a 3--8\% cleanup), regardless of
    mechanism (weak soft push / strong soft push / population-level
    decorrelation) or $\lambda$ (10$\times$ change bought nothing --- the
    penalty saturates).
  \item Every pre-registered bar MISSED: headline ($K{=}32,h{=}4 \geq
    0.5$) $\to$ measured $0.014$--$0.026$ ($\sim$20--25$\times$ below);
    minimum publishable ($K{=}16,h{=}4 \geq 0.8$) $\to$ measured
    $0.465$--$0.516$ ($\sim$1.6--1.7$\times$ above achieved).
  \item $h{=}1$ no-sacrifice guard SATISFIED everywhere --- the
    interventions do what they promise, just not enough of it.
\end{itemize}

\subsection{The proportionality reading: real, but geometrically shrinking}
\begin{itemize}
  \item Gram cleanup at $K{=}32$: 3--8\% real (every one of 9 seeds below
    baseline band). Recovery gain shrinks at every hop: $h{=}2$: $+0.07$
    to $+0.10$; $h{=}3$: $+0.04$ to $+0.05$; $h{=}4$: only $+0.005$ to
    $+0.017$.
  \item Exactly the shape the $\varepsilon^h$ compounding law
    (\S\ref{sec:phenomenon}) predicts --- a modest, real reduction in
    per-binding $\varepsilon$ produces a shrinking-in-reverse benefit as
    $h$ grows.
\end{itemize}

\subsection{Approach angles 4/5 (architectural robustness 2$\times$2 at
the 14M LM rung): not a qk-norm artifact; gating shows a
trend-not-confirmed amplification}
% Grounding: EXPERIMENT_LOG.md 2026-07-09 "ATTRACTOR-ROBUSTNESS 2x2"
% entries (build audit with the corrected same-corpus noise floors; n=1
% screening harvest; n=3 escalation harvest). Archives:
% experiment-runs/2026-07-09_attrrob_2x2_harvest/ and
% experiment-runs/2026-07-09_attrrob_2x2_escalation_harvest/.
% Note the pre-registered escalation rule and its outcome: the n=1
% trigger fired on the gating axis (+6.16 = 2.75 sigma) but the n=3
% escalation did NOT confirm (+4.31 = 1.92 sigma < the 2-sigma bar
% 4.489) -- any use of these numbers must carry the
% trend-not-confirmed framing, never an amplification claim.
\begin{itemize}
  \item Setting shift, stated explicitly: unlike angles 1--3 (train-time
    soft pushes on the synthetic task above), angles 4/5 are an
    architectural-robustness 2$\times$2 (qk-l2-norm $\times$
    Gated-DeltaNet-style decay) at the 14M real-data LM rung of the
    scale ladder, read with the same-corpus gram-deviation probe
    against corrected same-corpus cross-seed noise floors
    (openr1-mix-ext $2.244$, wikitext-mix-ext $2.217$); an $n{=}1$
    screening plus a pre-registered $n{=}3$ escalation on the one axis
    whose screening delta exceeded the $2\sigma$ trigger.
  \item The attractor PERSISTS in all four cells: same-corpus
    gram-deviation $18.2$--$25.2$, every cell $\geq 8\sigma$ above zero
    deviation in floor units --- neither axis removes it.
  \item \textbf{qk-norm EXONERATED ($n{=}3$-solid).} Every run in this
    program already used the community-standard qk-l2-norm mitigation;
    turning it OFF is a within-noise null: $\Delta = -0.103 =
    0.05\sigma$ vs.\ the corrected same-corpus floor. The attractor is
    cross-key POPULATION geometry, not the single-vector
    eigenvalue-stability axis qk-norm conditions --- the direct data
    answer to ``is this just the known qk-norm issue in disguise.''
  \item \textbf{Gating: a direction-consistent amplification trend that
    did NOT meet the pre-registered confirmation bar --- reported as a
    trend, never as a claim.} The $n{=}1$ screening fired the
    escalation trigger ($+6.16 = 2.75\sigma$); at $n{=}3$ the effect
    shrank to $+4.31 = 1.92\sigma$, below the pre-registered $2\sigma$
    bar ($4.489$), with all three paired seeds positive (exploratory
    Welch $p{=}0.062$). Directionally, gated arms carry MORE deviation
    while training to LOWER loss in every seed ($3.515$ vs.\ $3.682$
    final train loss) --- a healthier-looking checkpoint carrying the
    pathology no better (trend-level: worse), consistent with this
    section's one-attractor reading and in no tension with angles
    1--3.
  \item Consequence for \S\ref{sec:discussion}'s limitations: this
    PARTIALLY discharges the single-architecture-family item ---
    qk-norm is exonerated at $n{=}3$ and the Gated-DeltaNet-style
    decay direction now has a trend-level answer at 14M; DeltaProduct
    and production-scale gated configs remain untested.
\end{itemize}

\subsection{Honest conclusion (pre-registered negative branch, not a failed pilot)}
\begin{itemize}
  \item Verbatim framing to carry into the paper: ``a dominant mechanism
    whose matching fix does NOT move the frontier is itself a publishable
    finding.''
  \item Demonstration-claim tier verdict: FALSE for all three arms, stated
    plainly. Weaker true claim earned: a genuine, causal
    (premise-conditional) but small write-geometry cleanup, recovery
    moving proportionally in the predicted direction.
  \item F-geo-1's key-Gram number is compromised as premise-evidence
    (directly optimized by $L_{\mathrm{orth}}$) --- the verdict rests on
    behavioral $\mathrm{rec@0.9}$ bars only for those two arms. F-geo-2
    (ZCA) has no such carve-out and lands in the same basin anyway ---
    itself evidence the carve-out isn't hiding a different outcome.
\end{itemize}
