\section{Results: The Fix Demonstration}
\label{sec:results}

% Full grounding: EXPERIMENT_LOG.md "F-geo-3 WAVE VERDICT" 2026-07-04 +
% the completed K=32 n_iter=20 escalation cells (same archive directory).
% ENDING RESOLVED — no open numbers remain in this section.

\subsection{The pre-spend gate}
\begin{itemize}
  \item Predicted $K16\ h4{=}1.00$, $K32\ h4{=}0.77$ from measured
    cross-episode drift ($0.94$ at $K{=}16$, $0.90$ at $K{=}32$) ---
    training stabilizes keys relative to the untrained-probe reference,
    resolving the attack's cross-episode-drift risk favorably enough to
    authorize launch. (The gate was registered on $K{=}16$ only; the
    $K{=}32$ prediction was explicitly non-gating --- load-bearing for
    the pre-mortem answer on simulator calibration.)
\end{itemize}

\subsection{K=16: minimum-publishable headline --- HIT (Fig.~7 left)}
\begin{itemize}
  \item 3/3 admissible seeds. $h{=}4$: $0.95$--$1.00$ vs.\ bar $\geq 0.8$,
    vs.\ baseline $0.42$--$0.47$. $h{=}7$: $0.55$--$0.67$ vs.\ baseline
    $0.07$--$0.10$. $h{=}1$ no-sacrifice: $1.0$ everywhere.
\end{itemize}

\subsection{K=32: $\sim$45$\times$, fully admissible, bar narrowly missed
--- and the miss is attributed (Fig.~7 right)}
\begin{itemize}
  \item \textbf{Move 1 --- admissibility CLEARED.} The original
    \texttt{n\_iter=12} cells were 0/3 admissible solely on the
    Newton-Schulz fallback criterion (56/20{,}000 steps $= 0.28\%$ eigh
    fallback --- disqualifying under item 2 of the substitute stack, per
    pre-registration). The audit-verified \texttt{n\_iter=20} escalation
    completed: ALL 3 seeds fully admissible --- ZERO fallback steps,
    every substitute-stack criterion passed. The improvement is claimable
    at full evidentiary tier: $h{=}4$: $0.009 \to 0.390/0.416/0.504$
    ($\sim$45$\times$); ID $h{=}2$: $1.0$ vs.\ baseline $0.26$; $h{=}3$:
    $0.87$--$0.95$ vs.\ $0.05$. Admission-stack difference drops to a
    footnote (both stacks + realized pass rates still named).
  \item \textbf{Move 2 --- headline bar ($\geq 0.5$) NOT met; honest
    narrow miss, reported exactly.} $h{=}4 = 0.504/0.416/0.390$, mean
    $0.44$; 1/3 seeds at the line. Never rounded up, never restated as
    ``approximately met.''
  \item \textbf{Move 3 --- the miss is ATTRIBUTED.} Trained cross-episode
    key drift at $K{=}32$ measured $0.9037 < 0.95$ = the pre-registered
    HIGH band (pinned before the wave ran) $\Rightarrow$ outcome F:
    ``stable-not-just-orthogonal geometry is the bottleneck'' --- the
    mechanism the adversarial round (finding F1) named and measured in
    advance. Within-episode orthogonality is delivered ($\mathrm{resid}
    \approx 0$ by construction); cross-episode key IDENTITY is the
    residual requirement (per-identity $v_{\mathrm{eff}}$ cannot chase an
    episode-conditional key). Follow-on --- cross-episode key
    anchoring/stabilization (the direction named in the design's own
    outcome-F row) --- \textbf{has since run to completion across four
    waves; see the dedicated sub-section below, not future work.}
  \item \textbf{Fallback-irrelevance finding (report alongside).} The
    \texttt{n\_iter=20} cells (zero fallbacks) are numerically IDENTICAL
    to the \texttt{n\_iter=12} cells (56 fallbacks): $0.504/0.416/0.390$
    vs.\ $0.39$--$0.50$. The 56 ``corrupted'' steps changed nothing ---
    the residual gap is drift, not solver noise. This both validates the
    admission criterion as conservative (not explanatory) and rules out
    solver health as a competing account of the bar miss.
  \item Frame sentence for the section (and the paper): \emph{the fix
    moves the frontier decisively at $K{=}16$ (bar hit 3/3), moves it
    $\sim$45$\times$ at $K{=}32$ without clearing the pre-registered bar,
    and the residual gap is attributed to a pre-measured second
    mechanism} --- a complete causal chain, no loose ends.
\end{itemize}

\subsection{What does NOT move: $h{=}21$}
\begin{itemize}
  \item Literal-depth composition STILL COLLAPSES under F-geo-3 ---
    orthogonality fixes write-time interference between bindings, not the
    separate iteration-compounding mechanism
    (\S\ref{sec:phenomenon}'s depth-amplification law).
  \item Name explicitly as the fix's own scope boundary, not a gap in the
    evidence.
\end{itemize}

\subsection{The key-anchoring program's results (completed, Fig.~10,
Fig.~11) --- not a forward-reference}
\begin{itemize}
  \item This sub-section reports, in order, the four things the
    outcome-F follow-on program (\S\ref{sec:the-fix}) established. It
    closes what an earlier stage of this project left as its single most
    open thread; see \S\ref{sec:the-fix} for the full arc and gating
    history and Fig.~10/Fig.~11 for the figures.
  \item \textbf{(1) The behavioral gain, independently replicated.} $K{=}32$
    $h{=}4$ recovery lifts from a freshly-measured baseline $0.4105$ to
    $0.61$--$0.71$ across 9 seed-runs, 3 distinct seed sets, 2 architecture
    variants (global-$\lambda$ candidate (d), per-entity-$\lambda$
    candidate (d$'$)) --- upgraded from ``reproduced'' to ``independently
    replicated'' because 6 of the 9 runs used genuinely fresh seeds never
    drawn anywhere in this program before.
  \item \textbf{(2) The mechanism rejection (Outcome C), confirmed twice.}
    A per-entity engagement instrument reads $<$14--41\% engagement at
    every $K{=}32$ leg, both architecture variants, immune to the specific
    objections (anchor-norm assumption, $\lambda$-degeneracy,
    power-manufactured engagement) that forced a rescue attempt's own
    rejection.
  \item \textbf{(3) The confirmed explanation (constancy suffices, by
    ablation).} A frozen, never-trained anchor table (random or
    frame-potential init) at the same blend weight matches or exceeds the
    learned table's own gain ($0.67$--$0.76$ vs.\ $0.61$--$0.71$); a
    negative-median $r_e$ reading in the pure-random arm confirms the
    engagement instrument correctly reports ``nothing to align to.'' This
    is CONFIRMED-BY-ABLATION: a falsifiable prediction was registered
    before the confirming data existed, and it held.
  \item \textbf{(4) The measured capacity boundary.} Extending the
    identical design to $K{=}48$ ($K/d{=}0.75$): fully admissible (3/3),
    both named explanatory channels (drift-space closure $\sim$81\% of the
    gap; value-Gram relief $0.626\times$ the reference) measurably active,
    bar still missed 0/3 seeds (mean $0.0215$ vs.\ $0.024434$). The
    capacity curve completes as a cliff ($\sim$1.00 $\to$ $\sim$0.65 $\to$
    $\sim$0.02 at $K/d{=}0.25/0.5/0.75$), not a smooth decline, while the
    theoretical ceiling declines only gently over the same range.
  \item \textbf{(5) The boundary, located, then dissolved, then its
    leading confound exonerated (Fig.~cliff, Fig.~dose) --- three further
    waves completed after (4).} A dedicated localization wave fits the
    cliff's midpoint to $x_0{=}0.5455$ (95\% seed-level bootstrap CI
    $[0.5385,0.5513]$, width $0.0127$, $w{=}0.0597$) at
    $d_{\mathrm{state}}{=}64$, a $\sim$20$\times$ tightening over the
    prior 0.25-wide two-anchor bracket, from six points including four new
    intermediate loads ($K{=}34,38,42,46$). Re-running the identical
    $K/d$ window at $d_{\mathrm{state}}{=}128$ ($K{=}68,76,84,92$) shows
    $h4{=}1.0$ at all four loads, all seeds, 12/12 cells --- the
    transition is not shifted, it exits the measured window entirely
    (CONFIRM-SHIFTED, strong form). The leading confound between the two
    scales --- anchor-table coherence, forced upward by the Welch bound
    only when entity count exceeds $d_{\mathrm{state}}$ --- is tested
    directly: holding $d_{\mathrm{state}}{=}128$, $K{=}68$ fixed and
    injecting controlled, frozen coherence up to $0.40$ (exceeding
    $d{=}64$'s own realized training band, $0.373$--$0.385$) under a
    concentrated rank-4 structure produces $h4{=}1.0$ at every dose,
    10/10 cells, no exception --- the strongest exoneration this design
    can produce for that structure. Neither of the two most naturally
    testable single-variable accounts ($K/d$ ratio; scalar rank-4
    coherence) is sufficient; overlap structure (diffuse injection,
    PI-gated) and raw state capacity remain live, unadjudicated
    candidates.
  \item \textbf{(6) A fourth completed wave tests the raw-state-capacity
    candidate directly and finds SUPER-LINEAR, not ratio-invariant,
    growth.} Re-fitting the identical localization procedure at
    $d_{\mathrm{state}}{=}80$ and $d_{\mathrm{state}}{=}96$ locates
    $d{=}80$'s transition at $x_0{=}0.6779$ (95\% CI $[0.6683,0.6867]$,
    clean fit, $n{=}5$ seeds/$K$) --- entirely excluding the
    pre-registered ratio-invariant band $[0.4745,0.6165]$ centered on
    $d{=}64$'s own $x_0{=}0.5455$. \textbf{At $d_{\mathrm{state}}{=}96$, a
    wider grid ($K{=}72,78,84,90$) initially flagged 11 of 12 new cells
    inadmissible; an independently-audited repro instrument diagnosed a
    tolerance miscalibration in the convergence gate (all 4,608 flagged
    episodes resolve by $n_{\mathrm{iter}}{\leq}28$ against a production
    $n_{\mathrm{iter}}{=}20$), and the admission flag alone --- never the
    $h4$ measurement --- was recalibrated, unlocking the full grid at
    zero new GPU spend.} The re-fit on the complete, unlocked grid
    ($K{\in}\{69,72,78,84,90\}$, $n{=}3$/$K$) finds NO cliff anywhere
    through $K/d{=}0.9375$, but a non-monotonic curve
    ($h4$ means $0.9592/0.9216/0.9326/0.9581/1.0000$: a dip at
    $K{=}72/78$, partial recovery at $K{=}84$, an exact ceiling at
    $K{=}90$) that the sigmoid fit cannot localize --- 100\% of 4,000
    bootstrap resamples land at the fit's own degenerate parameter
    bounds. This rules out fixed-ratio invariance as cleanly as the
    $d{=}128$ wave ruled out ``the cliff never moves''; neither the
    absolute-slack band ($[0.718,0.739]$) nor the power-law band
    ($[0.768,0.837]$) is confirmed or excluded --- the precise functional
    form, and the exact $d_{\mathrm{state}}{=}96$ transition point, remain
    open, now for a scatter/power reason rather than a data-availability
    one. Seed escalation at the noisiest group ($K{=}72$, per-seed range
    $0.8426$--$0.9904$) is the registered next step, not launched by this
    wave.
  \item State plainly, as the paper's own honesty discipline requires: the
    behavioral gain is real and its cause is now known (constancy,
    confirmed by ablation); the originally-hypothesized entity-alignment
    mechanism is rejected; the fix's regime is bounded at
    $K/d{\leq}0.5$; that boundary is now precisely located and shown to
    dissolve with scale; the leading single-variable account of that
    dissolution is directly tested and exonerated, not merely assumed;
    and the raw-state-capacity candidate left open by that exoneration is
    itself now directly tested, ruling out simple ratio-invariance while
    leaving its exact functional form for a not-yet-launched follow-up.
    No unexplained residual remains on this thread except the two
    explicitly named, not-yet-fully-adjudicated mechanistic candidates.
\end{itemize}
