\section{Discussion and Limitations}
\label{sec:discussion}

% Full content: NARRATIVE.md \S6. State every item undiluted --- this is
% the section a reviewer checks first for overclaiming.

\begin{enumerate}
  \item \textbf{Small scale.} $d_{\mathrm{model}}$ in the low hundreds,
    $d_{\mathrm{state}}{=}64$, sub-1M--low-single-digit-M parameters (the
    core causal chapters). \textbf{Partial evidence, not a resolution:} a
    controlled 14M$\to$98M$\to$392M$\to$1.31B scaling ladder (Tier 2,
    geometry-only, a clean 4-point read on a single held-fixed corpus mix,
    Fig.~9 Panel A) finds the write-geometry attractor does NOT dissolve
    with scale --- it WORSENS monotonically through the full ladder
    (span-fraction $0.248 \to 0.344 \to 0.389 \to 0.455$), a
    \textbf{pure-scale} claim, now COMPLETE at four points rather than an
    open 3-point trend. \textbf{Disclosure, stated plainly:} the
    1.31B-parameter rung self-terminated at its own internal timeout at
    step $155{,}081/155{,}028$ of a planned $183{,}105$ ($\approx$84.7\%
    of the token-matched budget) --- a clean, designed shutdown (loss
    curves clean throughout, zero skipped steps), not a crash, and not
    recoverable by resuming (no optimizer-state checkpointing existed for
    this run). The harvested reading is neutralized against this shortfall
    by a dedicated plateau check at four late checkpoints
    (steps $130\text{k}$--$155\text{k}$): span-fraction is flat, mildly
    \emph{declining} over that window ($0.4584 \to 0.4554$, $-0.003$),
    linearly extrapolating to $\approx$0.452 at the full $183{,}105$-step
    budget --- two orders of magnitude smaller than the $+0.066$
    rung-2$\to$rung-3 increment it would be correcting, and in the
    direction that would slightly shrink, not grow, the reading. Harvesting
    at the 155k checkpoint (rather than a $\approx$144 GPU-h cold restart,
    for which no state exists to resume from) was the adjudicated choice
    on this basis. This item's core claim (untested at production scale
    for compositional recovery, not just geometry) stands regardless of
    that reading.
  \item \textbf{Single architecture family.} Vanilla, single-head,
    single-layer DeltaNet only. Gated DeltaNet (the actual Qwen3-Next
    production config), DeltaProduct's multi-Householder generalization,
    untested. $H{=}1$/single-layer was a deliberate gate-simplicity
    decision, not evidence of $H{=}1$-specificity (ReserveMH's
    generality result was run on the SYNTHETIC task only).
  \item \textbf{$h{=}21$ literal-depth decay remains unexplained and
    unfixed.} F-geo-3 demonstrably does not touch it --- a named,
    scoped-out mechanism (iteration-compounding), not folded into ``the
    fix works.''
  \item \textbf{The $K{=}32$ headline bar is not met, and its named
    follow-on has now run to completion with a fully resolved, if more
    complex than hoped, result.} The fix is fully admissible at BOTH $K$
    ($K{=}16$ 3/3; $K{=}32$ 3/3 on the \texttt{n\_iter=20} escalation,
    zero fallback steps), but the pre-registered $\geq 0.5$ headline bar
    at $K{=}32$ is narrowly missed: $h{=}4 = 0.504/0.416/0.390$, mean
    $0.44$, 1/3 seeds at the line. The miss is attributed to outcome F
    (trained cross-episode key drift $0.9037$ in the pre-registered HIGH
    band). \textbf{The learned anchor-table follow-on designed to close
    that residual (\S\ref{sec:the-fix}) has now run to completion across
    four waves,} not future work: it lifts $K{=}32$ $h{=}4$ recovery from
    a freshly-measured $0.4105$ baseline to $0.61$--$0.71$ (independently
    replicated, 9 seed-runs across 3 seed sets, 2 architecture variants);
    its own per-entity engagement instrument reads under 14--41\%
    engagement at every leg, confirming Outcome C (``mechanism not
    engaged'') at increasing rigor across two waves, immune to a rejected
    rescue attempt. \textbf{The actual explanation was then found and
    confirmed by ablation:} a frozen, never-trained anchor table (random
    or frame-potential init) at the same blend weight matches or exceeds
    the learned table's own gain --- constancy in the blend arithmetic,
    not entity-specific learned content, is the active ingredient; SGD's
    only demonstrated role is tuning the blend weight. \textbf{The fix's
    operating regime is now measured, not open-ended:} a capacity-curve
    extension to $K{=}48$ ($K/d{=}0.75$) shows the identical fix fully
    admissible, both of its own named explanatory channels measurably
    active, and the bar still missed 0/3 seeds --- the curve is a cliff
    ($\sim$1.00 $\to$ $\sim$0.65 $\to$ $\sim$0.02), not a smooth decline.
    State plainly: the behavioral $K{=}32$ $h{=}4$ gain is real and its
    cause is now known (constancy, confirmed by ablation); the
    entity-alignment hypothesis originally proposed for it is rejected;
    and the fix's regime is bounded at $K/d{\leq}0.5$. This item, open at
    an earlier stage of this project, is now closed.
  \item \textbf{The $K\approx d/2$ structural boundary is now a LOCATED
    cliff (Fig.~cliff), and its own dissolution and leading confound are
    both now measured, not merely observed.} The completed capacity-curve
    extension (item 4 above) turned this into a measured three-point curve
    at $d_{\mathrm{state}}{=}64$; a dedicated localization wave has since
    sharpened this to a fitted transition midpoint $x_0{=}0.5455$ (95\%
    seed-level bootstrap CI $[0.5385,0.5513]$, width $0.0127$,
    $w{=}0.0597$), a $\sim$20$\times$ tightening of the prior 0.25-wide
    bracket, fit to six points including four new intermediate loads
    ($K{=}34,38,42,46$). \textbf{The cliff is dimension-dependent, not a
    universal property of $K/d$:} re-running the identical $K/d$ window
    at $d_{\mathrm{state}}{=}128$ ($K{=}68,76,84,92$, the same ratios)
    shows $h{=}4$ at $1.0$ for all four loads, all seeds, $12/12$ cells ---
    the transition does not merely shift, it exits the measured window
    entirely (CONFIRM-SHIFTED, strong form). \textbf{The leading
    single-variable account of that dissolution --- anchor-table
    coherence, forced upward by the Welch bound whenever $n{>}d$ (as at
    $d{=}64$) and removed when $n{\leq}d$ (as at $d{=}128$) --- is directly
    tested and EXONERATED for a concentrated (rank-4) injection
    structure.} Holding $d_{\mathrm{state}}{=}128$, $K{=}68$ fixed and
    injecting controlled, frozen coherence up to $0.40$ (exceeding
    $d{=}64$'s own realized final-checkpoint training band,
    $0.373$--$0.385$) produces $h4{=}1.0$ at every dose, 10/10 cells, no
    exception (Fig.~dose). Neither of the two most naturally testable
    scalar accounts (K/d ratio; scalar rank-4 coherence) is sufficient on
    its own. \textbf{Two candidates survive, adjudicated by neither wave:}
    overlap STRUCTURE (a diffuse, \texttt{subspace\_rank=48} injection is
    a registered, PI-gated, not-yet-run co-primary arm --- this is not
    silently dropped, it is the paper's own open follow-on) and ABSOLUTE
    STATE CAPACITY ($d_{\mathrm{state}}$ grew $4\times$ while $K$ only grew
    $\sim$2$\times$ across the two waves, a confound no wave here holds
    fixed). \textbf{A dedicated follow-up wave has since tested the
    ABSOLUTE-CAPACITY candidate directly, by re-fitting the transition at
    two further $d_{\mathrm{state}}$'s, and REFUTES pure ratio-invariance:
    the transition's own load tolerance grows SUPER-LINEARLY in $d$, not
    at a fixed $K/d$ ratio.} Against a pre-registered ratio-invariant band
    $[0.4745, 0.6165]$ (centered on $d{=}64$'s own $x_0{=}0.5455$), the
    identical fitting procedure at $d_{\mathrm{state}}{=}80$ locates the
    transition at $x_0{=}0.6779$ (95\% CI $[0.6683,0.6867]$, clean,
    non-degenerate fit, $n{=}5$ seeds/$K$ after a seed escalation tightened
    an initial $n{=}3$ read of $x_0{=}0.6756$) --- entirely above the band,
    a real, disclosed exclusion. \textbf{A wider $d_{\mathrm{state}}{=}96$
    grid ($K{=}72,78,84,90$) was then run to attempt the same localization,
    and 11 of the 12 new cells were flagged \emph{inadmissible} by the
    standard convergence gate --- a wide-grid instrument-rigor episode
    with its own resolution, disclosed in full rather than silently
    dropped.} A dedicated, independently-audited repro instrument diagnosed
    the flag as a TOLERANCE MISCALIBRATION in the admission gate itself,
    not a real capacity wall: every diagnostic precondition passed cleanly
    (reconstruction gate, zero pool-membership violations across 4,608
    episodes, 36 reproduced events, zero live/offline disagreements), and
    the verdict-bearing test found \textbf{all 4,608 flagged episodes
    resolve by $n_{\mathrm{iter}}{\leq}28$} against a production setting of
    $n_{\mathrm{iter}}{=}20$ --- including the worst-case late-training
    population, where the near-miss residuals actually WORSEN over
    training (13--27/128 anomalous rows at an early checkpoint, maxima
    $0.07$--$0.43$, growing to 128/128 anomalous rows with maxima up to
    $1.43$ by the final checkpoint) yet still resolve fully within the
    modest iteration bump; a byte-identical determinism replay (12/12)
    independently ruled out a nondeterminism explanation. On this evidence
    the admission FLAG alone --- never the $h4$ measurement, which was
    always a valid training-time reading --- was recalibrated for the 11
    cells sharing the repro's own verified failure signature, unlocking
    the full grid at zero new GPU spend. \textbf{The re-fit on the
    complete, unlocked $d{=}96$ grid ($K{\in}\{69,72,78,84,90\}$,
    $n{=}3$ seeds/$K$) finds NO cliff anywhere in the tested window (through
    $K/d{=}0.9375$), but the curve is non-monotonic, not flat-at-ceiling:}
    per-$K$ means $0.9592/0.9216/0.9326/0.9581/1.0000$ dip through
    $K{=}72/78$, partially recover at $K{=}84$, and hit an exact ceiling
    ($1.0000$, all 3 seeds) at $K{=}90$, the widest load tested. The
    sigmoid fit is degenerate at its own parameter bounds ($x_0{=}0.9$,
    $L{=}1.2$, both pinned at their upper bound), and every one of 4,000
    bootstrap resamples lands in the same degenerate regime (100\%) ---
    \textbf{$x_0(96)$ does not localize, from genuine per-seed scatter
    rather than a data gap} (the unlock closed the data gap cleanly, 5/5
    $K$'s populated at $n{=}3$). Of the two simplest rival functional
    forms, an absolute-slack account ($x_0 \propto d_{\mathrm{state}}{-}\text{const}$,
    band $[0.718,0.739]$) and a power-law account (band $[0.768,0.837]$),
    \textbf{neither is confirmed or excluded by this wave} --- the
    scaling-law progression is $x_0{=}0.5455$ ($d{=}64$) $\to$
    $x_0{=}0.6779$ ($d{=}80$) $\to$ unresolved ($d{=}96$), for a different
    reason than the earlier data-gap failure: the data now exists, but is
    too noisy at $n{=}3$ seeds/$K$ for a monotonic-sigmoid model to
    resolve (the noisiest group, $K{=}72$, spans $0.8426$--$0.9904$, a
    range wider than the entire gap between the two rival bands). Seed
    escalation at $K{=}72$ is the registered next step, not launched by
    this wave. The open question is
    therefore sharper than before: not ``does
    it break past $d/2$'' (yes, precisely located) nor ``does the break
    persist at larger $d$'' (no, it dissolves) but ``what mechanism
    distinguishes a table forced non-orthogonal by the Welch bound from
    one that is not, if not scalar coherence magnitude alone.'' The
    design's own named, not-yet-tested candidate is value-crowding (more
    entities packed into the same fixed value dimensionality reducing
    per-entity value-recovery headroom independent of key-side
    conditioning), which would require a fresh design to test, alongside
    the structure/capacity candidates above.
  \item \textbf{Scale-transfer to a full pretrained LLM checkpoint is now
    PARTIALLY measured, and the fix mechanism's transfer to free-running
    text is separately tested and blocked.} Wave 2 establishes the
    truncation-damage phenomenon generalizes to naturalistic LM
    pretraining at small scale (descriptive+interventional tier only).
    Beyond that: (a) \textbf{the controlled 14M$\to$98M$\to$392M
    $\to$1.31B parameter ladder is now COMPLETE, a positive headline
    result in its own right}: on a single held-fixed data mix, the
    write-geometry attractor does not dissolve --- it WORSENS
    monotonically across all four points (span-fraction $0.248 \to 0.344
    \to 0.389 \to 0.455$), a \textbf{pure-scale} finding (Fig.~9 Panel A),
    confirming the failure mode is real, persists (rather than being a
    small-model artifact scale would eventually wash out), and is now
    precisely characterized through 1.31B parameters --- not yet extended
    to compositional recovery at scale. The fourth rung's own training run
    self-terminated at $\approx$84.7\% of its token-matched budget (a
    disclosed, designed timeout, not a crash); a dedicated plateau check
    at the final four logged checkpoints shows the reading flat-to-mildly-
    declining ($-0.003$ span-fraction over the last $25$k steps, against a
    $+0.066$ rung-2$\to$rung-3 increment), so the harvested value is not
    sensitive to the shortfall. (b) a much larger version
    of the SAME non-orthonormal write-geometry signature is measurable
    directly in production-scale pretrained fixed-state models (RWKV-7
    1.5B, Falcon-Mamba-7B; Fig.~9 Panel B); (c) but a matched
    no-fixed-state negative control (Qwen2.5-1.5B) shows the SAME
    magnitude, so (b) is explicitly NOT attributable to the delta-rule
    write mechanism specifically at this measurement tier; and (d) a
    companion attempt to install the SAME fix mechanism inside a
    free-running-text LM (Track B) is blocked before any interaction
    claim can be tested --- a duplicate-key numerical-stability smoke
    overshoots its bar by 63$\times$ with a probative positive control,
    compounding the original $\beta$-uniformity gate refusal; the
    selectivity-only readout that DID complete is INCONCLUSIVE (a split
    verdict across corpora). Net honest statement: the phenomenon
    persists where it can be controlled for and now scales purely with
    parameter count (a), a phenomenologically similar signature exists at
    production scale (b) but causal attribution there remains open and
    confounded (c), and the fix itself has not yet been shown to transfer
    to a free-running-text setting at all (d).
  \item \textbf{The admission-stack asymmetry, now footnote-level.}
    Because F-geo-3 makes the standard instruments tautological, a
    substitute pre-registered criterion applies to its seeds. With every
    substitute criterion passed at both $K$ (including zero fallback
    steps on the escalation cells), this drops from headline asterisk to
    footnote --- but still name both stacks and both realized pass rates
    wherever geo3 and non-geo3 seed counts appear side by side, because
    the two stacks filter different failure modes and their relative
    selectivity was never empirically equated.
  \item \textbf{The 2$\times$2's ``stable'' cells lean on existence-tier
    or bundled arms --- no longer an open dependency, a closed, mixed
    answer.} The 2$\times$2 (Fig.~5) that demonstrates orthogonality and
    stability are jointly necessary uses i-strong (pool-restricted to
    $32{\leq}d_{\mathrm{state}}$) for the ``both together'' cell and
    frozen-embedding arms for the ``stable, not orthogonal'' cells. The
    key-anchoring program (\S\ref{sec:the-fix}) was built specifically to
    test whether a full-vocabulary, gradient-discovered anchor mechanism
    can reach the ``jointly sufficient'' regime, and it has run to
    completion across four waves with a final, closed answer: the wave
    produces a real $K{=}32$ $h{=}4$ behavioral gain
    ($0.4105 \to 0.61$--$0.71$) but NOT via the hypothesized
    entity-alignment channel (Outcome C, confirmed twice); the gain's
    actual cause is confirmed by ablation to be constancy in the blend
    arithmetic, not learned content; and this alternate mechanism's own
    regime is bounded (works at $K/d{\leq}0.5$, fails at $K/d{=}0.75$).
    So: the 2$\times$2's ``jointly sufficient'' corner still rests on the
    hand-built i-strong pin for a \emph{causal} demonstration at $K{=}32$;
    the \emph{practical}, SGD-adjacent route tested to reach that corner
    does not use the hypothesized mechanism, works via a simpler
    (constancy-only) route instead, and that alternate route's own
    working regime is now known. A closed, three-part answer, stated as
    such.
  \item \textbf{The key-anchoring program is now fully complete, and its
    complete arc is part of this paper's evidence base.} Every claim
    above about ``the residual mechanism'' (outcome F, the 2$\times$2)
    still rests on the outcome-F attribution and the 2$\times$2, not on
    the anchoring program's numbers, for the causal MECHANISM claim about
    why F-geo-3 misses its $K{=}32$ bar. What the anchoring program
    contributes, now completely: a real, independently-replicated
    $K{=}32$ $h{=}4$ behavioral improvement; a rejected entity-alignment
    hypothesis (Outcome C, confirmed twice); a confirmed alternate
    explanation (constancy suffices, by ablation); a measured capacity
    boundary ($K/d{\leq}0.5$ works, $K/d{=}0.75$ does not); and a
    demonstrated instance of this project's own gating discipline both
    catching an attempted post-hoc rescue of a null result AND
    pre-registering and then confirming a falsifiable account of a
    surprising positive result. Report all of this, not just the
    flattering parts.
  \item \textbf{An attempt to connect this paper's causal capacity and
    geometry findings to natural in-context multi-hop reasoning ran into
    an instrument-validity problem, caught by the instrument's own
    pre-registered gates rather than silently reported as a finding ---
    itself evidence for, not against, this project's gating discipline.}
    A held-out-hop probe applied directly to this paper's own checkpoints
    (the frozen-bias intervention arms as the geometry leg, the
    14M--392M scale ladder as the scale leg) returned
    $\mathrm{recovered\_frac@0.9}{=}0.0$ at every one of 78 cells and
    every tested hop depth: the probe's own pre-registered Stage-0 sanity
    floor and alignment premises fail categorically, not marginally,
    routing the result to PROBE-INVALID rather than a licensed REFUTE
    (the design's own explicit rule: an instrument that cannot clear its
    own sanity floor does not get to report a negative result). A
    follow-up isolated whether this was a surface-form artifact of the
    original query-marker template: two structurally different
    natural-language reformulations --- one dropping the reserved marker
    entirely, one substituting an order-of-magnitude more corpus-common
    relation-verb family --- were run through the identical gate and
    failed IDENTICALLY (categorical failure at all 4 gate cells, the same
    shape as the original 78-cell result). \textbf{Two structurally
    different query formats failing the same way is the single most
    informative null this instrument can produce: it exonerates surface
    form as the cause and mechanically promotes a task-familiarization
    instrument (continued training on the probe's own query distribution
    before re-evaluating) as the next, already-designed step.} Realized
    cost of both probe-invalid harvests combined was small (${\approx}0.30$
    GPU-h against a ${\approx}24.2$ GPU-h ceiling); a disclosed process gap
    let the first harvest run to completion past its own already-failed
    gate (fixed and verified holding on the second harvest, which halted
    correctly before any full-grid spend). \textbf{The promoted
    task-familiarization instrument has since run to completion and closes
    the arc: a THIRD, structurally different instrument, at a third
    training regime, lands on the identical categorical floor.} Continuing
    training on the probe's own bind/query structure for 5,000 steps (6
    cells: 2 corpora $\times$ 3 seeds) and re-evaluating the SAME
    $d_{\mathrm{state}}$-space geometric readout at 5 checkpoints each
    returns $\mathrm{recovered\_frac@0.9}{=}0.0000$ at every one of 30/30
    (cell, checkpoint) readings --- the identical floor the marker and
    natural-language templates found, now after genuine task-directed
    training rather than zero-shot. \textbf{A real dissociation, built into
    the design deliberately, is worth reporting alongside the null:} a
    second, independent readout computed through a completely different
    pathway (vocab-space cross-entropy through the LM head, $L_{\mathrm{query}}$,
    never trained on the geometric quantity the gate measures) falls
    substantially and uniformly across all 6 cells (21.8--46.4\%, mean
    35.9\%) over the same 5,000 steps the geometric readout stays flat at
    exactly 0.0000 --- partial, genuine task engagement co-occurring with a
    categorically invalid geometric readout, not a training failure. The
    pre-registered $<$50\% pin for licensing the strong ``readout invalid
    despite complete task learning'' claim is not met in any of the 6
    cells (closest miss 46.4\%), so this is reported as
    PARTIAL-TASK-LEARNING-BELOW-PIN, not the strong claim. \textbf{This
    dissociation motivated promoting the vocab-space readout itself to a
    primary, independent behavioral-contrast instrument (no geometric
    readout at all)}, run as a fourth wave on the SAME familiarized
    checkpoints (12 new cells: $\{\mathrm{per\_token},\mathrm{global}\}
    \times$ 2 corpora $\times$ 3 seeds, plus the 6 reused OFF cells as
    baseline). Applying the pre-registered killer-prediction rule at all 4
    (corpus, arm) contrasts $\times$ 5 checkpoints: 3 of 4 contrasts are
    UNRESOLVED --- an underpowered null at $n{=}3$ seeds, exactly the
    outcome a pre-registered power check flagged as the single most likely
    result (the detectable $|\Delta|$ floor, $\approx$1.5--1.7 loss units,
    is the same order of magnitude as the OFF arm's own entire 5,000-step
    familiarization effect, $\approx$1.69). The 4th contrast
    (wikitext$\times$per\_token) is determinate and TRANSIENT:
    $\Delta(K{=}32,c{=}2500){=}{-}0.4999$, 95\% CI $[-0.6241,-0.3758]$ ---
    the arm's loss is HIGHER, not lower, than the untreated baseline at
    that checkpoint, resolving back toward the baseline by $c{=}5000$
    ($\Delta{=}{-}0.7949$, CI straddles zero) --- a non-durable
    training-dynamics artifact per the pre-registered rule, and in the
    HURTS, not helps, direction. Every determinate reading found across all
    20 primary checkpoints was negative-direction; none was positive.
    \textbf{Because the tested intervention is the arm's entire causal
    package} (the original pretraining-era weight divergence plus the same
    frozen-bias blend continuing to fire on every familiarization step,
    inseparable in what these checkpoints are, not an isolated
    familiarization-only contrast), this bounds rather than resolves the
    keystone causal question: does frozen-bias key-geometry stabilization
    change in-context multi-hop acquisition? Neither a PERSISTENT/
    LATE-EMERGENT reading (the only registered path to ``yes'') nor a
    CONVERGED-EQUIVALENT reading (the only registered path to a confident
    ``no'') occurred at any of the 4 contrasts --- this is a
    measurement-power result, not a mechanism result. \textbf{Net, honest
    statement:} the $d_{\mathrm{state}}$-space geometric readout construct
    is dead across all three surface forms and training regimes tested
    (marker zero-shot, natural-language zero-shot, task-familiarized) ---
    still routed to PROBE-INVALID, never REFUTE, per this project's own
    standing rule; separately, the promoted vocab-space behavioral
    instrument bounds any causal effect of the arm's whole causal package
    at under 1.5--1.7 loss units at this seed budget, with the one
    determinate signal found being real but transient and pointed in the
    hurts, not helps, direction. No claim about in-context reasoning,
    positive or negative, is licensed by any of the four harvests.
\end{enumerate}
