\section{Related Work}
\label{sec:related}

\paragraph{Fast-weight and linear-attention architectures.}
\citet{katharopoulos2020linear} identify linear attention with a
recurrent state; \citet{schlag2021linear} connect it to fast weight
programmers and introduce the delta-rule update this testbed trains;
\citet{yang2024deltanet} give the parallel kernel used here. DeltaProduct
\citep{siems2025deltaproduct}, negative-eigenvalue variants
\citep{grazzi2025negative}, and gated delta rules
\citep{yang2024gateddelta} extend the family's expressivity, and
matrix-state RNNs push it to language-model scale \citep{mishra2026m2rnn};
state-space models make the same constant-memory trade
\citep{gu2023mamba}, with known expressivity limits
\citep{merrill2024illusion}. Titans \citep{behrouz2025titans} and
Infini-attention \citep{munkhdalai2024infiniattention} add a bounded,
test-time-updated memory module alongside or in place of attention,
targeting the same fixed-memory budget this paper measures. None of
these works locates a trained capacity transition or measures its
movement with state dimension, which is this paper's object.

\paragraph{Capacity of associative memories.} Classical correlation-matrix
memories establish the $K \lesssim d$ regime for linear reads
\citep{kohonen1972correlation, anderson1972simple};
\citet{barnfield2026sharp} prove sharp thresholds for linear associative
recall across retrieval rules. \citet{nichani2025factual} show that under
argmax decoding a rank-one state recovers on the order of $d$
associations: the readout, not the state alone, sets the accounting. This
paper's contribution is empirical and complementary: a located frontier
for a \emph{trained} delta-rule memory under an exact continuous readout,
the regime where within-state capacity is load-bearing.

\paragraph{Memory management for long-context inference.} KV-cache
eviction and compression manage a memory footprint that grows with
context \citep{xiao2024streamingllm, zhang2023h2o}, and recall-throughput
trade-offs in efficient architectures are measured by
\citet{arora2023zoology} and \citet{arora2024based}, whose
linear-attention models are architecturally adjacent to the single-head
DeltaNet measured here. This paper addresses the fixed-size endpoint of
that trade-off: not how much a growing cache can be shrunk, but how much
a state of constant byte size can be trusted to hold.

\paragraph{Rank in trained linear-attention states.}
\citet{nazari2026rank} and \citet{sun2026staterank} describe effective-rank
dynamics of linear-attention states in trained LLMs. Those are
descriptive accounts of what rank trained states occupy; the present
work measures where task recall collapses as bound load approaches the
state's dimension, with pre-registered fits and a causal dose-response
test on the leading confound.
