\section{Limitations}
\label{sec:limitations}

\paragraph{Scope.} All results use a single-head, single-layer DeltaNet,
one synthetic 107-entity recall grammar, sub-million-parameter models, and
$d_{\mathrm{state}} \in \{64, 80, 96, 128\}$
($d_{\mathrm{state}} = 32$ is excluded by a kernel-level failure in the
underlying library, measured independently of this program). No claim
transfers to production-scale gated variants
\citep{siems2025deltaproduct, grazzi2025negative}, to matrix-state models
at language-model scale \citep{mishra2026m2rnn}, or to natural language.
The per-byte conservation reading of Table~\ref{tab:bytes} rests on two
located points plus one in-window bound; it is a measured regularity over
$d_{\mathrm{state}} \in [64, 96]$, not a law. Every cell is trained for a
fixed budget of $20{,}000$ steps; whether the located transitions move
under a substantially larger training budget is untested, so the reported
frontier is a trained-at-fixed-budget capacity frontier, not a storage
bound.

\paragraph{The held-out axis is hop depth, not entity identity.} Every
$h4$ measurement in this paper draws its queried entities from the same
107-entity pool used in training; the held-out axis is hop depth alone
(Section~\ref{sec:testbed}). A separate disjoint-entity-pool diagnostic,
recorded at every checkpoint under a pre-registered scope limit (the
anchoring mechanism claims no held-out-entity gains, and held-out
entities bypass the anchor blend by construction), collapses at hop
depth 2 and above: recovered fraction at threshold $0.9$ is $1.0$ at
hop 1 in all 24 archived cells behind
Figure~\ref{fig:cliff}, and at hop depth 2 it reads exactly $0.0$ in 23
of those 24 cells (the remaining cell reads $0.0002$) and $0.0$ in all
24 at hop depth 3,
while the same checkpoints' trained-pool readouts stay at ceiling at
hops 1 and 2 and remain well above the held-out collapse at hop 3,
where they decline under the highest tested loads (as low as $0.56$
at $d_{\mathrm{state}}=64$, $K{=}46$). % evidence: C17
The located frontier and its scaling with $d_{\mathrm{state}}$ are
therefore claims about capacity for entities the model has trained on,
not about generalization to unfamiliar entities; the latter is a
substantially harder, unresolved question on this testbed.

\paragraph{The frontier's functional form is open.} Four dimensions
constrain the growth of $x_0(d)$ but do not pin its form: the two located
fits refute ratio invariance, yet the absolute-slack and power-law
accounts both remain live at $d_{\mathrm{state}}=96$, where the fit is
$100\%$ bootstrap-degenerate from real per-seed scatter. % evidence: C11
Discriminating them requires either variance reduction at the noisiest
loads or a window that brackets the $d_{\mathrm{state}}=96$ transition
from above, and a registered analysis found the present grid underpowered
for that discrimination; we therefore make no functional-form claim.
% evidence: C13

\paragraph{Near-ceiling fine structure at $d_{\mathrm{state}}=96$ is
instrument-limited.} The unlocked grid's non-monotonic dip-and-recovery
pattern, including the exact-ceiling reading at $K{=}90$, sits close to
the admission instrument's own resolution. A registered two-cell
fresh-seed diagnostic aimed at this fine structure routed to its
pre-registered degenerate outcome: neither cell produced an admissible
reading at the recalibrated gate, and the failing admissibility leg
differed from the one the original grid failed, so the diagnostic's
adjudication triggers were never licensed to fire. % evidence: C14
One of the two cells, at $K{=}90$, read $h4 = 0.9725$ before its
inadmissibility was determined; the archived $K{=}90$ ceiling
($h4 = 1.0000$ in each of three seeds, Table~\ref{tab:bytes}) did not
replicate at this fresh seed. % evidence: C14
Per the diagnostic's own pre-registered routing this reading cannot be
promoted to a positive finding, because the cell is inadmissible, but we
disclose it here as a caveat that discounts confidence in the archived
$K{=}90$ reading.
We therefore scope the $d_{\mathrm{state}}=96$ claim strictly as the
absence of a monotonic collapse through $K/d = 0.9375$, and we treat the
per-load ordering near ceiling as unresolved. The admission gate episode
itself (Section~\ref{sec:frontier}) cuts both ways: it shows the
measurement pipeline can manufacture an apparent collapse, and it is
disclosed precisely so that the reported transitions are read against
that risk; the located $d_{\mathrm{state}} \in \{64, 80\}$ fits pass
their gates without any recalibration.

\paragraph{The dose test bounds the tested variable only.} The
exoneration of Section~\ref{sec:dose} covers scalar coherence dose under
two frozen injection structures (rank-4 and rank-48). It does not cover
coherence that co-evolves with training: a non-frozen
$d_{\mathrm{state}}=128$ reference point whose table coherence drifted to
$[0.127, 0.202]$ under ordinary training also recovers at $h4 = 1.0$,
% evidence: C3
but such a point confounds the coherence level with how the level was
reached, and is reported as context, not as part of the controlled sweep.
Mechanistic accounts below the level of the disclosed comparison axes
(for example, optimization dynamics through the larger eval-time
Newton--Schulz block) are untested here.

\paragraph{One architecture family.} Whether the same three-part
structure, a sharp transition, its dissolution with scale, and a
coherence exoneration, appears in other fast-weight update rules or
state-space families is untested; the measurement recipe (pre-registered
logistic localization, degenerate-fit disclosure, saturation checks,
admission gates) is the transferable artifact.
