\section{Task, Model, and Instrument}
\label{sec:setup}

\textbf{Task.} A word $w = g_{i_1} \cdots g_{i_L}$ is drawn as a random
walk on $G$'s Cayley graph over a symmetric generating set
($L \sim \mathrm{U}\{1,\dots,8\}$) and must map to
$\rho_G(g_{i_1} \cdots g_{i_L})$, the composed product's image under a
pinned orthogonal reference representation of dimension $\dmin(G)$. The
five groups are $S_3, S_4, A_5, S_5, A_6$ with $\dmin = 2, 3, 3, 4, 5$
(realizations in Appendix~\ref{app:groups}); $S_3$ and $S_4$ are
solvable, the rest are not.

\textbf{Model.} A Transformer encoder with $\dstate$ learned row-reader
queries writes each word into a single state
$Z(w) \in \R^{\dstate \times \dstate}$, with $\dstate = \dmin + 2$ per
group so that over-recruitment is expressible. The loss is cosine
distance to the block-diagonal embedding of the reference
($\rho_G(\cdot) \oplus I_2$ in the observational sweep,
$\rho_G(\cdot) \oplus 0$ in the causal wave); the readout is fixed, with
no learned weights that could launder rank, and the decoder reads only
$Z$, verified by a gradient blank-out test with a planted-leak negative
control. Per-group step budgets (8k--40k, recorded in the design
document's pre-registration) were pinned by convergence bars before the
decisional sweep ran.

\textbf{Instrument.} The learned state is defined only up to scale and
an orthogonal change of basis (a Schur intertwiner):
$Z(w) \approx c\,Q\,[\rho_G(w) \oplus \cdot\,]\,Q^{\top}$. We estimate
the model's own dominant $\dmin$-subspace $U$ from the SVD of the
\emph{centered} covariance of $Z(w)$ over held-out words, measure
\textbf{restricted effective rank}, the entropy effective rank of
$U^{\top} Z(w) U$, and score \textbf{degauged recovery}, cosine to the
reference after fitting $(c, Q)$ on a fitting split and evaluating on a
disjoint split, reported as $\recninety$ (the fraction of eval words
above cosine 0.9). Rank is invariant under the fitted gauge, so a
rank-deficient state cannot be degauged into a full-rank one; centering
is load-bearing (Appendix~\ref{app:instrument}). The observational band
$[0.7, 1.3] \cdot \dmin$, the equivalence margin and sample size, the
force-rank grid and its recovery bar, and every falsifier were fixed
before the decisive runs.
