\section{Related Work and Limitations}
\label{sec:related}

\citet{merrill2024illusion}, \citet{grazzi2025negative}, and
\citet{siems2025deltaproduct} frame state tracking by what recurrent
architectures \emph{can} express on exactly these word problems, the
solvability axis; this work measures what trained states \emph{do}
converge to, and finds the learned dimension sorts by representation
theory instead. \citet{nazari2026rank} and \citet{sun2026staterank}
measure state-rank dynamics in pretrained linear-attention models
observationally, with no algebraic ground truth to converge to and no
causal leg. \citet{nichani2025factual} prove associative-memory capacity
under argmax decoding, the loophole our exact-continuous-recovery
scoring closes; \citet{mishra2026m2rnn} train matrix-state RNNs on $S_3$
composition but measure no state rank and run no rank intervention.
\citet{chughtai2023toy} reverse-engineer which representation-theoretic
structure networks recruit on finite-group composition tasks; this work
asks a narrower question on the same substrate, at what dimension, with
a pre-registered causal intervention absent there.
\citet{huh2024platonic} propose that models trained on different tasks
converge to a shared representation; the present law is a narrower,
algebra-driven instance of that convergence.

\textbf{Limitations.} All results are on sub-1M-parameter synthetic
models under a deliberate single-state bottleneck; nothing here is a
claim about pretrained networks, and the family is five groups with
$\dmin \in \{2,\dots,5\}$ at $\dstate = \dmin + 2$. Causal cells are
single-seed except $S_3$ (four seeds), with the necessity zeros exact
wherever seeds exist; two groups at the shortest pinned budget ($S_3$,
$S_5$) show soft convergence and are also the two with the largest
observational deviation and the one marginal causal cell, a
pre-registered rather than post hoc budget schedule but a caveat on
cross-group comparison strength. The observational readout partially
favors rank $\approx \dmin$ by construction, which is why it was
pre-registered as corroborating and the causal wave carries the claim.
