\documentclass{article} % COLM 2026 template (official kit, tag 2026) -- same
% vendored kit as papers/reasoning-null-moss/ and papers/measurement-ws/
% (this program's other MOSS-class honest-negative-with-mechanism papers).
\usepackage[submission]{colm2026_conference}

\usepackage{microtype}
\usepackage{hyperref}
\usepackage{url}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{lineno}

\definecolor{darkblue}{rgb}{0, 0, 0.5}
\hypersetup{colorlinks=true, citecolor=darkblue, linkcolor=darkblue, urlcolor=darkblue}

\title{The K-Axis Closes at 32 -- And Why: Diagnosing a Spectrum-Blind
Write Objective in a Fast-Weight Composition Reader}

% Anonymized for double-blind review (MOSS @ COLM 2026); the submission
% option suppresses this block regardless.
\author{Anonymous authors}

\begin{document}

\ifcolmsubmission
\linenumbers
\fi

\maketitle

\input{sections/00_abstract}
\input{sections/01_introduction}
\input{sections/02_background}
\input{sections/03_win}
\input{sections/04_wall}
\input{sections/05_diagnosis}
\input{sections/06_pending}
\input{sections/07_related}
\input{sections/08_discussion}

\bibliography{refs}
\bibliographystyle{colm2026_conference}

\appendix

\section{Registered Gates and Bands, Reproduced}
\label{app:gates}

This appendix reproduces, in compact form, the pre-registered gates and
verdict maps the main text relies on. Each was fixed in the program's
design registry (\texttt{matrix-thinking/NOVEL\_ARCH\_WATERFALL.md},
\texttt{matrix-thinking/NCR\_ORTHO\_WRITE.md}) before the corresponding
data existed.

\textbf{Axis-A separation bands (\S\ref{sec:win}).} Per arm, a numeric
band at the separation depth $h^\ast$ on median $\mathrm{recovered\_frac@0.9}$
across seeds: HOLD $\geq 0.9$; DEGRADED $\in (0.5, 0.9)$; FAIL $\leq 0.5$.
The baseline axis is the best-performing comparison-of-record arm. A
per-arm-pair cell (NCR band, best-baseline band) maps to WIN / SEP-PARTIAL
/ TIE / LOSE per a $3\times3$ table; two per-$K$ labels combine via a
second table (WIN+WIN$\to$WIN; WIN+SEP-PARTIAL or WIN+TIE$\to$WIN-PARTIAL;
any pair containing LOSE$\to$LOSE).
% evidence: NOVEL_ARCH_WATERFALL.md S3.2a
Axis B (sequential cost) is claimable only if Axis A $\geq$ TIE, and is a
WIN at $\geq 10\times$ wall-clock advantage at the deepest cost-probe
point. Axis C (spectral predictability) is a WIN if the locked prediction
tracks measured cosine within 0.05 through $h \leq 509$ for $\geq 3/5$
seeds.
% evidence: NOVEL_ARCH_WATERFALL.md S3.2/S3.2a

\textbf{\S11 Gate 1/Gate 2 (\S\ref{sec:wall}).} GATE 1 (convergence), per
$K$ over 4 seeds: in-distribution ($h{=}1,2,3$, min) $\mathrm{recovered\_frac@0.9}
\geq 0.9$ AND mean $A_{\mathrm{eff\_rank}} \geq 0.9K$
$\to$ CONVERGED; in-dist $\in [0.5,0.9) \to$ PARTIAL; in-dist $< 0.5
\to$ DEAD. Per-$K$: CONVERGED-ROBUST ($\geq 3/4$ CONVERGED) /
CONVERGED-PARTIAL (1--2/4) / TRAINABILITY-DEAD (0/4). GATE 2 (far-depth),
scored only on CONVERGED cells: $\mathrm{recovered\_frac@0.9}$ at the
cell's own $h^\ast = 8K{-}3$, banded HOLD ($\geq 0.9$) / DEGRADED
($0.5$--$0.9$) / FAIL ($\leq 0.5$).
% evidence: NOVEL_ARCH_WATERFALL.md S11
The K=32 $d(K)$-grid stopping rule: REOPENS/CONVERGES-ONLY both require
CONVERGED-ROBUST at some $d \in \{K{+}1, 1.25K, 1.5K\}$; if none clears
it, the K-axis is CLOSED-AT-THIS-K and the next rung is BLOCKED.
% evidence: NOVEL_ARCH_WATERFALL.md S11.5
The budget-rescue ANOMALY check fires if $\delta$ is non-monotonic across
the $1\times{\to}2\times{\to}4\times$ trajectory in $\geq 3/4$ seeds (or
either of two other named triggers); a fired ANOMALY pre-empts the
mechanical verdict map and reports the trajectory numbers-only.
% evidence: NOVEL_ARCH_WATERFALL.md S11.6

\textbf{Ortho-write WIN/PARTIAL/NULL/FAIL (\S\ref{sec:pending}).} Gate-0
precondition (both arms): in-dist $\mathrm{recovered\_frac@0.9} \geq 0.9$
and mean $A_{\mathrm{eff\_rank}} \geq 0.9K$; a cell failing Gate-0 is DEAD
and unscored on far depth. Part A (primary, $K{=}32$, realistic $h^\ast=40$):
WIN requires ortho-write median $\mathrm{recovered\_frac@0.9}$ at
$h^\ast{=}40 \geq 0.9$ across converged seeds AND the pinned free-write
baseline $< 0.5$ at $h{=}40$, with the mechanistic signature corroborating
(departure-from-normality $\leq 0.02$, condition number $\lesssim 2$,
$\min|\lambda|/c^\ast \geq 0.9$); PARTIAL requires median
$\mathrm{recovered\_frac@0.9} \geq 0.9$ at $h \in \{20, 29\}$ but $< 0.9$
at $h{=}40$; NULL is median $< 0.9$ at every depth $\geq 12$ while Gate-0
still passes; FAIL is Gate-0 DEAD in $\geq 3/4$ seeds. Part B (structured-
operator discriminator) uses the identical bands with path length $L$ in
place of $h$ and $L^\ast{=}32$.
% evidence: NCR_ORTHO_WRITE.md S4

\section{Reproducibility}
\label{app:repro}

Every number in this paper is transcribed from the program's design
registry (\texttt{matrix-thinking/NOVEL\_ARCH\_WATERFALL.md},
\texttt{matrix-thinking/NCR\_ORTHO\_WRITE.md}), never from memory or
re-derivation, with the source section marked in a LaTeX comment beside
each claim in the source of this paper. Diagnosis scripts
(\texttt{k32\_spectral\_diag.py}, \texttt{k32\_counterfactual.py},
\texttt{k32\_nonnormal\_frontier.py}) are archived at
\texttt{experiment-runs/2026-07-16\_ncr\_ortho\_write/diagnosis/}. The
ortho-write fix experiment (\S\ref{sec:pending}) was, as of this draft, a
live pre-registered but not-yet-read GPU wave; this paper does not read
its results path and reports the pre-registered bands only, as pending.

\end{document}
