\begin{abstract}
Fixed-state sequence models bind in-context associations in a matrix-valued
fast-weight state; geometric observables have been proposed as lenses on
composition. We test whether this observable predicts, or causally
improves, in-context multi-hop composition in DeltaNet-family language
models spanning 14M to 1.31B parameters, and find a null bounded three
ways. First, a state-space composition readout, deployed three
structurally different ways, returned a recovered fraction of exactly zero
at 366 of 366 readings, % evidence: C1-phase1
gates failing at every cell. % evidence: C1-phase1
A pre-submission positive control caught a state-layout defect in the
readout; % evidence: C10-instrumentfix
after the fix, the 320-reading marker-template sub-grid no longer reads
zero, % evidence: C11-remetric
but two pre-registered correspondence nulls reproduce the signal at every
one of those readings: recovery there is indistinguishable from a broken
correspondence; the other two deployments keep their pre-fix zero
unconfirmed. % evidence: C12-nullvalidation
Second, a
vocabulary-space behavioral contrast bounds any causal effect of a
write-geometry intervention below its pre-registered three-seed detection
floor of 1.5 to 1.7 loss units; % evidence: C5-power
the sole reading meeting the pre-registered differential condition (one of
five that exclude zero across forty tests) is a transient, harm-direction
deviation. Third, a twelve-seed replication of the transient was
stopped by its pre-registered batch-effect gate (cohort variance ratio
4.47 against a heuristic cutoff of 4.0), % evidence: C7-replication
and the new cohort's interval spans zero. Each bound names the observation
that would overturn it. The result is a null with visible assumptions and
closed escape routes.
\end{abstract}
