\section{Introduction}

Linear-attention and delta-rule language models compress an entire context
into a fixed-size matrix state
\citep{schlag2021linear,yang2024parallelizing}. A companion research program
measured \emph{geometric} observables of the written state (e.g.\ span
fraction, effective key rank) across scales and under targeted
interventions, carrying an unexamined promise that write geometry matters
for in-context multi-hop composition. We test that promise and find a
null. A reader of a null has two standing worries, that the instrument
never worked and that the analysis could declare a null after seeing the
data. Three pre-registered bounds address both:

\begin{enumerate}[leftmargin=1.5em,itemsep=1pt,topsep=2pt]
\item \textbf{An instrument bound.} A state-space composition readout,
deployed three structurally different ways, returned a recovered fraction of
exactly zero at 366 of 366 readings under the pre-fix instrument,
% evidence: C1-phase1
its validity gates failing at every cell. A pre-submission positive
control on the readout's production path failed outright, exposing a
state-layout defect; % evidence: C10-instrumentfix
after the fix, the marker-template sub-grid (320 of the 366 readings) no
longer reads zero, % evidence: C11-remetric
but two pre-registered correspondence nulls reproduce the apparent signal
at every one of those readings; % evidence: C12-nullvalidation
the other two deployments were not independently re-verified
(Section~\ref{sec:geometric}). One reading clearing a null, or one premise
gate passing, would have voided this bound; none did.
\item \textbf{A power bound.} A vocabulary-space behavioral contrast, with
a three-seed detection floor of 1.5 to 1.7 loss units registered before
unblinding, % evidence: C5-power
leaves three of the four per-arm contrasts unresolved; the sole reading meeting the
pre-registered differential condition is a transient, harm-direction
deviation. % evidence: C6-transient
A determinate effect above the floor would have voided this bound.
\item \textbf{A replication bound.} A targeted twelve-seed extension of
that transient was stopped by its own pre-registered batch-effect gate
(a 4.47-fold cohort variance mismatch against a heuristic 4.0 cutoff),
% evidence: C7-replication
and the new cohort's interval spans zero. A clean pooled interval excluding
zero would have voided this bound.
\end{enumerate}

The contribution is the bounded null and the measurement discipline that
kept it interpretable.
