\section{Models, Task, and Instruments}
\label{sec:setting}

\textbf{Checkpoints and task.} All experiments probe DeltaNet-family
language models \citep{yang2024parallelizing} pretrained on a
reasoning-dense corpus mix (OpenR1-Math) and an encyclopedic mix
(WikiText-103, \citealp{merity2017pointer}). An \emph{intervention grid}
(14M parameters) varies a frozen-key-bias intervention on the fast-weight
write path in three validation-loss-matched arms: control, per-token, and
global. A \emph{scale ladder} spans 14M, 98M, 392M, and 1.31B parameters
(state dimension 64, 64, 128, 128; depth 2, 12, 16, 22).
% evidence: C0-ladder
Each probe episode binds $K$ entities into one directed cycle with
\textsc{bind} statements, then queries the entity $h \in \{1,\dots,4\}$
hops around the cycle; a full $K$-cycle prevents held-out hop depths from
collapsing modulo short cycle lengths. Load is $K$ relative to state
dimension ($K \in \{20, 32\}$ on the grid; near-capacity per rung).

\textbf{Instrument 1: a geometric readout.} For a layer's terminal state
$S_T$ and effective query $q_{\mathrm{eff}}$, the readout predicts the
$h$-hop target as $S_T^{\,h} q_{\mathrm{eff}}$ and scores \emph{exact
continuous recovery}: the fraction of queries reaching cosine at least 0.9
with the target value vector. Continuous recovery, rather than argmax over
candidates, follows \citet{nichani2024understanding}, under which argmax
decoding can succeed without the state carrying the claimed structure.
Registered validity gates accompany the readout: an $h{=}1$ sanity floor
and two alignment premises, (iii) query--key and (iv) key--value, each
requiring a same-entity median cosine above a shuffled-entity null's 95th
percentile (Appendix~\ref{app:gates}); gate failure routes to
probe-invalid, never to refutation.

\textbf{Instrument 2: a behavioral contrast.} A familiarization phase
trains checkpoints for 5{,}000 steps on the bind/query task mixed into the
pretraining corpus; the contrast is the held-out-episode query loss $L_q$
(a vocabulary-space cross-entropy through the language-model head,
machinery disjoint from Instrument 1), differenced control minus arm at
five checkpoints. A six-outcome trajectory classification and a
differential condition (the $K{=}32$ effect determinate while $K{=}20$ is
not) were fixed before unblinding. Every wave ran gate-first, with
verdicts routed mechanically (Appendix~\ref{app:gates}).
