\section{The Geometric Readout Never Fires}
\label{sec:geometric}

\begin{table}[t]
\centering
\footnotesize
\setlength{\tabcolsep}{3pt}
\begin{tabular}{@{}llll@{}}
\toprule
Wave (instrument) & Grid & Readings & Registered outcome \\
\midrule
1 zero-shot, marker$^{\ddagger}$ & 78 cells, 14M--392M & 312/312 $= 0.0$ & probe-invalid \\ % evidence: C1-phase1
2 zero-shot, 1.31B rung$^{\ddagger}$ & 2 cells & 8/8 $= 0.0$ & probe-invalid \\ % evidence: C2-rung3
1+2 fixed-lens re-metric & same 80 cells & 78/320 nonzero, $h{=}1{\leq}0.87$ & unadjudicated \\ % evidence: C11-remetric
1+2 correspondence nulls & same 80 cells & 0/320 clear either of 2 nulls & null-indistinguishable \\ % evidence: C12-nullvalidation
3 zero-shot, natural lang.$^{\S}$ & 4 gate cells & 16/16 $= 0.0$ & probe-invalid, refused \\ % evidence: C3-phase1b
4 familiarized gate$^{\S}$ & 6 cells $\times$ 5 ckpts & 30/30 $= 0.0$; $L_q$ falls & dissociation \\ % evidence: C4-dissoc
5 behavioral contrast & 18 cells, 4 contrasts$^{\dagger}$ & 3 unresolved, 1 transient & power-bounded \\ % evidence: C5-power
6 seed extension, $n{=}12$ & 18 pretrain cells & variance ratio 4.47 & batch-effect-flagged \\ % evidence: C7-replication
\bottomrule
\end{tabular}
\caption{The program at a glance. Waves 1--4 deploy the geometric readout
(recovered fraction at cosine $\geq 0.9$). Waves 5--6 deploy the
vocabulary-space contrast; the one determinate transient is in the harm
direction and its $n{=}12$ replication was stopped by the pre-registered
batch-effect gate.
$^{\dagger}$Per-arm re-derivation; corpus-level returned both unresolved.
$^{\ddagger}$Raw reading superseded below; independent premise failure
still routes to probe-invalid. $^{\S}$Identical pre-fix function, not
independently re-run (Appendix~\ref{app:gates}).}
\label{tab:program}
\end{table}

\textbf{The readout was defective, and the fix did not save it.} A
pre-submission positive control on the readout's production
state-extraction function failed outright (0 of 256 recovered at every
hop), exposing a $[K,V]$-vs-$[V,K]$ state-layout transpose; after a
one-line fix and an independent review, the same control recovers 1.0000
at every hop (Appendix~\ref{app:gates}). % evidence: C10-instrumentfix
Re-running the wave-1/wave-2 grid (80 cells, 320 readings) through the
corrected readout no longer returns zero: 78 readings are nonzero, one
$h{=}1$ cell reaching 0.87. % evidence: C11-remetric
Two pre-registered correspondence nulls -- a label-shuffle null, later
found vacuous at $h{=}1$ by construction, and a positional derangement
null with teeth at every hop -- both reproduce this apparent signal at
every one of the 320 readings (0/320 clear either bar), tracing it to
collapsed prediction directions scored against a near-collinear value
population (Appendix~\ref{app:gates}). % evidence: C12-nullvalidation
Bound 1's claim is that recovery is statistically indistinguishable from a
broken correspondence, not that it reads zero; the alignment-premise gates
(Figure~\ref{fig:gates}) fail at all 320 readings regardless, before and
after the fix. % evidence: C1-phase1, C2-rung3, C11-remetric

\textbf{Zero-shot, natural language (wave 3).} Two template families -- a
gift-verb family with the query marker dropped, and a corpus-native
succession-verb family -- fail identically: 16 of 16 readings at 0.0, the
$h{=}1$ gate false at all 4 cells, not a surface-form artifact.
% evidence: C3-phase1b

\textbf{Task-familiarized (wave 4).} The strongest-case test trains
control checkpoints on the task itself for 5{,}000 steps and applies the
gate at five checkpoints per cell. It fails at 30 of 30 readings, 0 of 512
queries recovered each time, % evidence: C4-dissoc
while the vocabulary-space query loss falls 21.8 to 46.4 percent (mean
35.9 percent) in 6 of 6 cells, from checkpoint 250 (first post-warmup) to
checkpoint 5{,}000 % evidence: C4-dissoc
(Figure~\ref{fig:dissoc}, Appendix~\ref{app:figs}): the models measurably
learn the task through one readout while the premise gates fail
throughout and the recovered-fraction floor -- a pre-fix reading,
Appendix~\ref{app:gates} -- stays at 0.0. The dissociation, on the
premise gates rather than the pre-fix floor, indicts the readout
construct, not the models (partial task learning, short of the registered
50-percent pin). Waves 3--4 share the pre-fix function and were not
independently re-run; their probe-invalid routing rests on the unaffected
premise gates alone, not the raw zero reading.
% evidence: C3-phase1b, C4-dissoc

\textbf{Gates with enforcing code paths.} Wave 1's chain computed its
abort gate but never acted on it, running the full grid past an
already-invalid probe; every later chain enforced its gate at the process
boundary, and waves 3 and 4 each refused launch on a real failing gate.
% evidence: C8-teeth
