\section{The Behavioral Contrast Is Power-Bounded}
\label{sec:behavioral}

Wave 5 promotes the vocabulary-space contrast to the primary instrument:
all three arms are familiarized under identical recipes (18 cells), and
the arm effect is the held-out query-loss difference, control minus arm,
so positive means the intervention helps.

\textbf{The registered power floor.} Before unblinding, the three-seed
detection floor was derived from the control cells' between-seed spread: a
$\sigma$ proxy of 0.43 to 0.48 loss units gives a 95-percent half-width of
roughly 1.5 to 1.7. % evidence: C5-power
For calibration, the entire familiarization effect is about 1.69 loss
units: % evidence: C5-power
the instrument detects only effects about as large as everything the model
learns in the phase, so the registration anticipated unresolved outcomes,
to be read as measurement bounds, never as evidence of no effect.

\textbf{Results.} The registered corpus-level classifier, which uses the
global arm as each corpus's representative, returns both corpora
unresolved; a disclosed build-time per-arm re-derivation (folded into the
analysis code, re-validated against the archived deltas;
Appendix~\ref{app:gates}) splits this into four contrasts, three
unresolved. % evidence: C5-power
The fourth (encyclopedic corpus, per-token arm) classifies as transient:
at checkpoint 2{,}500 the $K{=}32$ effect is determinate at $-0.500$,
interval $[-0.624, -0.376]$, while $K{=}20$ is not, and by checkpoint
5{,}000 it dissolves
% evidence: C6-transient
(Figure~\ref{fig:transient}, left, Appendix~\ref{app:figs}). A
held-out-hop secondary readout fires at the same cell and checkpoint, same
direction, % evidence: C6-transient
consistent with one coherent mid-training deviation.

\textbf{Sign discipline and multiplicity.} Of 40 uncorrected 95-percent
intervals in the primary table, three $K{=}32$ readings exclude zero, near
the two expected by chance, all in the harm direction (arm loss above
control); two share checkpoint 1{,}000 across different cells, consistent
with sampling noise (Appendix~\ref{app:gates}). Only the
checkpoint-2{,}500 cell also satisfies the pre-registered differential
condition and draws the held-out-hop corroboration; that compound
condition, not any single interval, is the one determinate signal.
% evidence: C6-transient
``The intervention matters'' ignores that transients were registered as
training-dynamics artifacts; ``no effect'' ignores the power floor.
