\section{The Replication Gate Refuses the Pool}
\label{sec:replication}

Wave 6 extends the transient's anchor cell (encyclopedic corpus, per-token
arm, $K{=}32$, checkpoint 2{,}500) to twelve seeds, with three integrity
mechanisms registered before launch: nine fresh pretraining seeds for both
arms (18 new cells); an archived-values loader with a runtime guard so the
original three seeds are never re-scored live (their deltas appear
byte-identical in the new harvest); % evidence: C7-replication
and a batch-effect gate requiring, before any pooling, cohort means within
twice the pooled standard error and a variance ratio below a pinned 4.0.

\textbf{The gate fired.} The archived cohort's between-seed standard
deviation (1.154) is 2.1 times the new cohort's (0.546): variance ratio
4.47, twelve percent past the cutoff, while the means agree comfortably
(shift 0.364 against a 1.382 threshold). % evidence: C7-replication
The registered routing reports the cohorts separately and classifies the
wave batch-effect-flagged, with no decision-grade pooled interval. The
archived cohort reproduces its original interval
$[-0.624, -0.376]$ by construction; the new nine-seed cohort's own
interval is $[-0.506, +0.357]$, spanning zero. % evidence: C7-replication
The naive twelve-seed pool, diagnostic only, would read
$[-0.509, +0.147]$ and would have classified the transient as refuted;
% evidence: C7-replication
the gate exists because pooling a high-variance three-seed cohort with a
lower-variance nine-seed cohort without a variance check assumes exactly
what a replication is supposed to test
(Figure~\ref{fig:transient}, right, Appendix~\ref{app:figs}).

\textbf{What stands.} The transient is neither confirmed nor refuted; it
keeps its original three-seed weight, no more. The evidence leans one way:
7 of 10 primary and 7 of 10 held-out (checkpoint, load) pairs are
computable, and every computable pooled interval contains zero.
% evidence: C7-replication
The variance mismatch is not established as a systematic between-wave
shift; its direction flips across checkpoints, most consistent with
small-$n$ variance imprecision in the archived cohort. A replication wave
whose own integrity gate can refuse produces a well-defined result even
without a verdict.
