\section{Scope of the Bounds}
\label{sec:scope}

\textbf{Bounded.} The readout construct, one layer's terminal state
applied $h$ times to a query under exact continuous recovery, carries no
signal in this checkpoint family at any tested scale, surface form, or
training regime. % evidence: C4-dissoc
Intervention effects beyond 1.5 to 1.7 loss units either way are excluded
at the registered confidence; % evidence: C5-power
the transient is bounded in duration, direction, and weight.
% evidence: C7-replication

\textbf{Not bounded.} The hypothesis survives below the power floor and
through mechanisms the readout cannot see, most plausibly cross-layer
hand-off (one layer resolves the first hop, a later layer the next); the
readout tests only single-layer self-iteration. A competing account is
expressivity: \citet{grazzi2024unlocking} prove positive-eigenvalue
DeltaNet cannot compose certain permutation groups, so the null may
partly reflect this variant's ceiling rather than a wrong observable;
wave 4's dissociation points toward the observable; and nothing here
transfers beyond the DeltaNet family or the bind/query grammar. Finally,
the readout's recovery math is validated only at unit level, never
end-to-end on a real forward pass with a known target, so a systematic
extraction or state-convention error is not fully excluded
(Appendix~\ref{app:gates}); a real-kernel positive control is the next
wave's priority.

\textbf{Lessons for small-scale practice.} A registered gate without an
enforcing code path is a sentence, not a safeguard (refusal worked twice
on real failures); % evidence: C8-teeth
routing fixed before unblinding kept zeros from becoming a refutation and
a transient from becoming a headline.
