\section{Introduction}
\label{sec:intro}

Continuous chain-of-thought (CoT) models replace explicit textual
reasoning steps with continuous latent tokens fed back into the
transformer's residual stream. COCONUT \citep{hao2024coconut} and
CODI \citep{shen2025codi} are representative instances: both
compress an explicit rationale into a small number of continuous
latent positions and decode the answer from the resulting state.
Theoretical work \citep{zhu2025superposition,gozeten2025cot2} argues
that these latents can hold multiple reasoning paths in
\emph{superposition}, so continuous CoT could explore a search tree
in parallel.

Recent empirical work has pushed back. \citet{rizvi2026illusion}
showed that a fine-tuned COCONUT model reaches 96.6\% on ProsQA
\emph{without} feeding back any latent tokens, against 99.0\% with
latents and 85.3\% for explicit CoT. In their fine-tuned setting,
the latent machinery is not doing work. They named this the
\emph{Illusion of Superposition}.

Matrix-valued latents \citep[e.g.,][]{shen2025codi} sharpen the
superposition hypothesis by introducing a single-sample structural
observable: rank. A $d\times d$ thought $\Z$ has a computable rank
via its SVD. If each singular direction encodes a separate reasoning
path, truncating $\Z$ to rank $k$ at inference should degrade
accuracy when the task needs more than $k$ paths. The rank-$k$
ablation curve is the natural probe.

\paragraph{Contributions.} We report four pieces of evidence that
this probe is uninformative in practice and give a
mechanism that explains why.

\begin{enumerate}
\setlength\itemsep{2pt}
\item \textbf{Three flat rank-$k$ curves} from the flatten-then-project
      readout on ProsQA across two distillation weights
      ($\gamma\!\in\!\{0,1\}$) and a multiplicative-thinker on/off
      ablation (\S\ref{sec:flat}). Range across
      $k\!\in\!\{1,2,4,8,16\}$ is $\leq 0.6$pp. A fourth flat curve
      on GSM8K-Aug is reported at a below-threshold ($6\%$) operating
      point and flagged as non-interpretable on its own.
\item \textbf{Three-seed accuracy-rank decoupling} (\S\ref{sec:flat}).
      Best ProsQA accuracy is $81.51\pm 1.2$pp while the final
      effective rank of $\Z$ spans $\{4, 12, 13\}$. The loss does
      not reward any particular rank.
\item \textbf{Four nonlinear-in-$\Z$ positive-control readouts also
      produce flat curves} (\S\ref{sec:pc}). A bilinear
      reparametrization, a bilinear-plus-GELU readout, an
      SVD-augmented readout feeding singular values through an MLP,
      and a quadratic readout in $\Z\Z^\top$ all yield flat rank-$k$
      curves (Spearman $p$-values $0.63, 0.14, 0.82, 0.46$).
      Readout linearity is a sufficient but not necessary condition:
      the matrix-bottleneck training objective itself produces
      rank-indifferent gradients through the full chain rule.
\item \textbf{A linear probe} on the matrix thought $\Z$ concatenated
      across six positions (1536 features) reaches AUC $0.673$ at
      predicting the ProsQA target class, below a raw pretrained
      GPT-2 hidden state at 768 features (AUC $0.846$,
      \S\ref{sec:flat}). The matrix bottleneck does not deliver
      target-predictive information commensurate with its feature
      dim.
\item \textbf{A negative control on vanilla GPT-2 SFT}
      (\S\ref{sec:pc-negctrl}). Running the rank-$k$ ablation on a
      model with no matrix bottleneck and no $\Z$ produces a flat
      curve at the same accuracy, and a random-$h$ sensitivity floor
      lands at the same accuracy. The rank-$k$ probe alone conflates
      rank-blindness with position-irrelevance; the matrix-CODI
      claim rests on the model-level seed-decoupling result, not on
      the probe's flat shape in isolation.
\end{enumerate}

Two February 2026 papers measure rank in \emph{linear-attention
fast-weight hidden states} \citep{nazari2026rank,anonymous2026staterank}
and describe rank as a structural invariant of trained states. The
contribution here is a \emph{mechanism} claim about the training
objective, and the four positive-control readouts are designed to
falsify that claim (\S\ref{sec:related}).

Collectively, these results indicate that rank-$k$ ablation is not a
valid measure of whether a matrix-valued latent holds multiple
reasoning paths under a CODI-style objective, and that the observed
rank of $\Z$ is, in our setting, vestigial.
