\section{Background}
\label{sec:background}

\paragraph{CODI distillation.} CODI \citep{shen2025codi} trains a
student to compress an explicit chain-of-thought into a fixed number
of continuous latent positions by matching a hidden-state target
from a teacher pass. The teacher pass consumes prompt $+$ CoT $+$
answer and produces a reference hidden state at a designated
colon-token position. The student pass consumes prompt $+\;n$ latent
positions $+$ answer, where each latent position is produced by
feeding the previous step's hidden state back as the next input
embedding. A hidden-state L1 loss (the \emph{distillation} loss)
aligns the student's state at the answer colon to the teacher's,
and a standard next-token cross-entropy loss trains the answer
prediction. The total loss is $\mathcal{L}=\gamma\mathcal{L}_{\text{kd}}+\mathcal{L}_{\text{ce}}$.

\paragraph{Why matrix latents.} Vector representations are flat: a hidden
state of dimension $D$ has $D$ scalar entries with no native notion of how
many independent components it superposes. Matrix-valued representations
have an additional structural observable: rank, the number of independent
directions used. Matrix-valued working memory has a long lineage in
neural networks --- fast-weight networks
\citep{schmidhuber1992fast,schlag2021linear}, linear attention
\citep{katharopoulos2020transformers}, the xLSTM family
\citep{beck2024xlstm}, and the Mamba/SSM family \citep{gu2023mamba} all
use matrix-shaped state --- and recent work measures the rank of these
states as a diagnostic of how the model uses its memory
\citep{nazari2026rank,anonymous2026staterank}. In all of those, the matrix
lives \emph{inside} a layer (an attention head's accumulator, an SSM's
recurrent state) and serves as recurrent memory. Matrix-CODI extends the
lineage to a different placement: the matrix sits on the explicit
chain-of-thought feedback path, where each latent reasoning position is a
$d \times d$ matrix rather than a vector. The rank of that matrix is a
candidate measure of how many reasoning paths the model holds in
superposition at that step. The hypothesis we test is whether rank, so
defined, behaves as that measure under CODI-style training.

\paragraph{Matrix bottleneck.} We extend CODI with a
\emph{matrix bottleneck} on the latent feedback path. Given the
previous latent hidden state $h\in\R^{D}$ (here $D=768$ for GPT-2
small):
\begin{align*}
\text{flat} &= W_{\text{up}}\,h,\qquad W_{\text{up}}\in\R^{d^2\times D}\\
\Z &= \text{reshape}(\text{flat};\,d,d)\in\R^{d\times d}\\
\Z &\leftarrow (I+\Delta(\Z))\,\Z\,(I+\Gamma(\Z)) \quad\text{(optional thinker)}\\
h_{\text{out}} &= \text{LayerNorm}(\phi(\Z)),
\end{align*}
where $\phi:\R^{d\times d}\to\R^{D}$ is the \emph{readout}. The
default readout is $\phi(\Z)=W_{\text{down}}\vecop(\Z)$, which we
call the \emph{flatten-then-project} readout. $W_{\text{down}}\in
\R^{D\times d^2}$. We use $d=16$ throughout.

\paragraph{Rank-$k$ ablation.} At inference, compute the SVD
$\Z=U\Sigma V^\top$ and replace $\Z$ by its rank-$k$ truncation
$\Z_k = U_{:,\,:k}\,\Sigma_{:k,\,:k}\,V_{:,\,:k}^\top$ before the
readout. If rank is functional, accuracy should drop as $k$ decreases
to 1. If rank is vestigial, the curve is flat.

\paragraph{Effective rank.} Because $\Z$ is a dense matrix of a
trained model, its numerical rank is always $d$. We report the
\emph{effective rank}
$\exp\left(-\sum_i \tilde\sigma_i \log \tilde\sigma_i\right)$ with
$\tilde\sigma_i=\sigma_i/\sum_j \sigma_j$, a smooth proxy used in the
spectral analysis literature. The ablation itself uses the hard
top-$k$ truncation.

\paragraph{ProsQA.} ProsQA \citep{hao2024coconut} is a synthetic
entailment task: a diamond-shaped directed acyclic graph of entities
and a question \emph{which leaf entity has property $P$?}. Each
problem has a unique positive answer and a single distractor. We use
the training split from
\texttt{facebookresearch/coconut}. The test set has 128 problems,
matching the evaluation protocol in the original CODI release.

\paragraph{Rank is not numerical rank.} Throughout, ``rank'' refers to
\emph{effective rank} (for training-curve reporting) and \emph{hard
SVD truncation rank} (for the rank-$k$ ablation). The dense-matrix
numerical rank of $\Z$ is always $d$ and does not change during
training; the question is whether the structural capacity offered by
rank $>1$ is functionally used by the model.
