\section{Related Work}
\label{sec:related}

\paragraph{Latent chain-of-thought.}
COCONUT \citep{hao2024coconut} replaces explicit rationales with
continuous latent tokens. CODI \citep{shen2025codi} compresses an
explicit CoT into latents via a self-distillation loss and matches
explicit CoT accuracy on GSM8K at GPT-2 scale. The Matrix-CODI
variant we study here inserts a $d\times d$ bottleneck on the
latent feedback path; our results are specifically about that
bottleneck and its readout.

\paragraph{Illusion of Superposition.}
\citet{rizvi2026illusion} show empirically that a fine-tuned
COCONUT model reaches $96.6\%$ on ProsQA \emph{without} latent
tokens, $99.0\%$ with them, and $85.3\%$ with explicit CoT. Their
evidence is behavioral and interpretability-based (logit lens,
entity-level probing). Our vanilla SFT baseline at GPT-2 small reaches $81.77\%$, roughly
$15$pp below their reported $96.6\%$; we did not close this gap. The
qualitative phenomenon they report --- fine-tuned latent CoT models
reaching comparable accuracy without their latents --- replicates at
our operating point: matrix-CODI at $82.03\%$ vs.\ pure SFT at
$81.77\%$ (gap $0.26$pp, within three-seed noise). Our contribution
relative to that work is a mechanism at the architecture level that
does not depend on matching their operating point: the
matrix-bottleneck training objective does not reward rank of the matrix
latent, and neither readout linearity nor readout nonlinearity changes
this. A rank-$k$ ablation on Rizvi-Martel's released checkpoint (code
at \url{github.com/michaelrizvi/coconut}) is a natural camera-ready
experiment that would test whether the same rank-blindness holds at the
$96.6\%$ ceiling.

\paragraph{SIM-CoT (ICLR 2026).}
\citet{shen2025simcot} diagnose latent CoT instability as
insufficient step-level supervision and propose injecting
per-step targets. Our diagnosis is different: the matrix-bottleneck training objective
produces rank-indifferent gradients, and our four positive-control
readouts adjudicate. If the SIM-CoT
diagnosis were the complete story, fixing the readout's Jacobian
would shift \emph{which} component of $\Z$ the gradient shapes; it
would not eliminate rank-dependence entirely. Our finding that all
four readouts (linear, nonlinear-via-GELU, SVD-augmented,
quadratic) produce flat rank-$k$ curves is consistent with both
mechanisms operating simultaneously. The two diagnoses are
compatible and non-redundant.

\paragraph{Reasoning by Superposition and CoT2.}
\citet{zhu2025superposition} prove that a two-layer transformer
with $D$ steps of continuous thought can solve directed graph
reachability, with each continuous thought encoding a parallel BFS
frontier. \citet{gozeten2025cot2} show similar parallel-exploration
behavior under a GRPO-style training regime. Both are theoretical
capacity results with small empirical demonstrations. We do not
contradict them: capacity and what-gets-learned are distinct, and
our result is specifically about what CODI distillation shapes, not
about whether transformers can in principle encode superposition.

\paragraph{February 2026 rank measurements (direct adjacency).}
Two February 2026 papers measure rank in trained transformer
states:

\citet{nazari2026rank} measure the effective rank of linear
attention hidden states and propose post-training rank pruning of
$K$ and $Q$ matrices. The \emph{object of study} is the
\emph{fast-weight matrix} inside a linear attention layer (a $d
\times d$ memory built inside the attention computation, not an
explicit CoT token). Their \emph{claim type} is descriptive --- they
observe rank structure, report a distribution, and prune.

\citet{anonymous2026staterank} (``State Rank Dynamics in Linear
Attention LLMs'') report a ``state-rank stratification'' phenomenon
during pretraining: linear-attention heads bifurcate into
persistently low-rank and persistently high-rank groups. Again the
object is the fast-weight memory, and the claim is descriptive of
what trained networks end up with.

Our paper differs from both on two axes:
(i) \emph{object of study} --- we measure rank in \emph{latent
thought matrices}, the explicit per-position $\Z$ tokens on the
feedback path of a matrix-CODI model, not a fast-weight memory
inside an attention layer; (ii) \emph{claim type} --- we make a
\emph{mechanism} claim about the training objective
(Proposition~\ref{prop:jac}) and we \emph{test it by falsification}
using four nonlinear-in-$\Z$ positive controls, which a descriptive
paper cannot do. We are not the first to empirically measure rank
in trained transformer states; those two papers do that for
linear-attention fast-weight states. We are the first (to our
knowledge) to diagnose rank-blindness as a property of the
\emph{readout-plus-objective} and to falsifiably test that
diagnosis with explicit nonlinear-in-$\Z$ readouts.

\paragraph{Dynamics within latent CoT.}
\citet{anonymous2026dynamics} run multiple intervention protocols on
latent CoT hidden states (zero, mean, step-wise mean, Gaussian noise)
and an early-stop decoding that truncates latent computation after step
$k$. Their early-stop decoding is a cousin of our rank-$k$ ablation on
a different axis --- step depth vs.\ spectral truncation. They find
step-wise causal structure, suggesting some latent positions carry
necessary information. Our rank-$k$ null across configurations,
combined with our linear-probe decay $\Z[1]\to\Z[5]$, is consistent
with their picture (early positions carry the information) while adding
a spectral-axis finding they do not measure.

\paragraph{Rank decay from depth.}
\citet{dong2021attention} showed that pure attention loses rank
doubly exponentially with depth. Their subject is the rank of
\emph{activations in a stack of attention layers}. Ours is the rank
of an \emph{explicit matrix latent} on a feedback path. The two
phenomena are distinct: rank-blindness in our setting persists with a
single bottleneck at each latent position, independently of stack
depth.

\paragraph{Implicit low-rank bias in matrix factorization.}
Gradient descent on matrix-factorization losses has an established
implicit bias toward low-rank solutions
\citep{gunasekar2017implicit,arora2019implicit,razin2020implicit},
and \citet{kobayashi2024weightdecay} show that weight decay in
particular induces low-rank attention products via an equivalence
with nuclear-norm regularization. These results concern the
optimizer's bias through the parameter space; our
Proposition~\ref{prop:jac} concerns the loss gradient through the
readout, and the two are compatible. Our three-seed spread in
effective rank of $\Z$ ($\{4,12,13\}$ under AdamW at weight decay
$0.01$) is consistent with an implicit low-rank attractor plus
seed-dependent convergence, and is inconsistent with strong low-rank
collapse (which would predict all three seeds at the same low rank).
A sharper treatment of which optimizer channel is shaping $\Z$'s
effective rank is left to future work.

\paragraph{Alternative substrates and methodological caveats for
ablation-based interpretability.}
\citet{wang2025latentvocab} propose that latent reasoning in LLMs
is a superposition in the \emph{vocabulary column space}, not in
the hidden-state basis --- a distinct substrate from the SVD
directions of $\Z$ our rank-$k$ ablation probes. If the relevant
superposition lives in the vocab-column subspace, rank-$k$
truncation on $\Z$ is not the right observable; our negative result
would not rule it out. \citet{fan2025rala} show that rank
augmentation is architecturally achievable in \emph{linear
attention} for vision via specific nonlinear interventions; our
``no nonlinear readout helps'' finding is evidence about the
readout-plus-objective family we tested and should not be read as
a universal claim about the design space. \citet{li2024optimal}
point out that zero/resample ablations --- of which our rank-$k$
truncation is an instance --- substantially overestimate component
importance relative to optimal ablation; that methodological
caveat cuts in our favor (optimal ablation would make our curves
flatter, not bumpier) but is the right citation for the probe we
use. The negative control in \S\ref{sec:pc-negctrl} is the empirical
counterpart to that critique applied to our own probe: running the
rank-$k$ ablation on a model with no matrix bottleneck reproduces a
flat curve, demonstrating directly that the probe alone does not
isolate rank-blindness from position-irrelevance.
