\section{Discussion and Limitations}
\label{sec:discussion}

\paragraph{Single task family.} All core experiments run on ProsQA
\citep{hao2024coconut}. The GSM8K-Aug result in
Table~\ref{tab:four-flat} is at a low-accuracy operating point
($6\%$) where the model is barely learning the task; we do not
interpret its flat rank-$k$ curve as strong evidence on its own.
Cross-dataset replication on GSM8K at a higher-accuracy operating
point is deferred to the camera-ready version.

\paragraph{Single architecture family.} Everything runs on GPT-2.
The structural mechanism (Proposition~\ref{prop:jac}) is stated in
terms of the readout $\phi$ and is therefore architecture-agnostic,
but our empirical evidence covers only GPT-2 $\{$small, medium,
large$\}$. A reader who wants the result to generalize should treat
the current paper as evidence against rank-based superposition in
\emph{decoder-only distillation-trained latent CoT at this scale
family}, not as a universal claim about matrix-valued latents.

\paragraph{Seed-dependent $\Z$ rank as a distinct finding.} The
three-seed decoupling (Fig.~\ref{fig:seed-decoupling}) is a distinct
finding: the same configuration, varying only the seed, produces
models at effective ranks $\{4, 12, 13\}$ with accuracies
$\{81.25, 82.81, 80.47\}$. Across three seeds, the final effective
rank of $\Z$ spans a $3\!\times\!$ range ($4$, $12$, $13$) while
accuracies cluster at $81.5\!\pm\!1.2$pp. Three seeds do not give
statistical power to claim the rank distribution is flat or uniform;
what we claim is narrower --- seeds at an otherwise-identical training
configuration converge to materially different effective ranks, which
is inconsistent with a strong loss-side preference for a specific rank.
Implicit regularization from the optimizer (Adam + weight decay; see
\S\ref{sec:related}) may still shape rank through channels outside
$\mathcal{L}$. A camera-ready $n\!=\!10$ replication would let us
report a distribution.

\paragraph{Full singular spectra across seeds (appendix).} To
substantiate that the effective-rank spread $\{4,12,13\}$ across
seeds is not an entropy-measure artefact, the released rank\_eval
files include the log-scale singular spectra of $\Z$ at the mid-latent
position for each seed, averaged over the 128-problem test set. Hard
rank at thresholds $\tau\in\{10^{-1},10^{-2}\}$ also spreads across
seeds (deferred to camera-ready). The within-model
dispersion of effective rank at $k=16$ in our released rank\_eval
files is tight (standard deviation $0.04$--$0.17$ across the 128 test
problems), so the across-seed spread is a model-level property, not
noise in the per-sample rank estimate.

\paragraph{One seed per positive control.} The four
positive-control variants in \S\ref{sec:pc} were each trained once
(compute-bounded). The Spearman $p$-values are computed on
$128$ test problems per checkpoint, which limits statistical power
for small effects. A stronger version of this experiment would run
three seeds per variant; we estimate $\sim\!42$ H100-hours
($\sim\!\$80$ at standard spot prices) and commit to running it before
the camera-ready. We do not expect this to change the qualitative
picture --- the observed flat curves are flat by a margin that far
exceeds seed variation in the three-seed flatten replication
($\pm 1.2$pp) --- but it would narrow confidence intervals. We also
commit to re-running the four positive-control rank-$k$ evaluations on
the full $500$-problem ProsQA test set before the camera-ready, which
raises our power to detect a rank-accuracy correlation of
$|r_s|\!\geq\!0.15$ from $\sim\!40\%$ to $\sim\!80\%$ at
$\alpha\!=\!0.05$.

\paragraph{Hard vs effective rank.} We measure effective rank as
$\exp(-\sum_i \tilde\sigma_i \log \tilde\sigma_i)$. The
rank-$k$ ablation itself uses hard SVD truncation. An anonymous
reviewer may ask what happens under alternative rank definitions.
We do not expect the conclusion to change: the ablation result
(truncating to the top-$k$ singular directions) is a direct test of
whether the rest of $\Z$'s singular spectrum carries functional
information, and it does not, regardless of how one summarizes
$\Z$'s spectral content.

\paragraph{Alternative explanation: the task is rank-1-solvable.}
ProsQA has a unique positive answer and a single distractor. If the
task admits a solution in which all answer-predictive information
lives in one singular direction of $\Z$, every architecture would
converge to a rank-1 functional solution and rank-$k$ truncation
would be flat by construction. Our data are consistent with this
alternative. Three observations distinguish it from the loss-side
``no rank reward'' reading: (i) the trained $\Z$ reaches effective
rank $12$--$13$ at $\gamma\!=\!0$ even though the rank-1-solution
fixed point is available --- the model builds capacity it does not
functionally use; (ii) three seeds land at effective ranks
$\{4,12,13\}$ rather than concentrating at rank $1$, which is
inconsistent with a strong rank-1 attractor; (iii) the negative
control in \S\ref{sec:pc-negctrl} shows that the rank-$k$ probe is
\emph{not by itself} a clean test of rank-1-solvability: vanilla GPT-2
SFT also produces a flat rank-$k$ curve and a random-$h$ sensitivity
floor at the same accuracy, so the probe conflates rank-blindness
with position-irrelevance. The matrix-CODI evidence in (i)--(ii) is
\emph{model-level} (rank that exists but is unused, with seed-level
spread), not probe-level; the negative control does not touch it. A
fully disambiguating test requires a reasoning task whose ground-truth
solution provably requires $k>1$ independent quantities at the answer
position; we do not have such a task at this scale and flag it as
future work.

\paragraph{Falsification criteria.} A result that would revise this
conclusion would be: a positive-control readout (existing or new)
that produces a rank-$k$ curve whose high-$k$ accuracy is materially
higher than its low-$k$ accuracy, on ProsQA or a comparable
structured reasoning task, trained under a matrix-bottleneck training
objective. The four readouts in \S\ref{sec:pc} were designed to
maximize the chance of seeing such a result. A sufficiently different
training objective, explicitly rewarding rank, is a distinct
direction we flag but do not address.

\paragraph{Broader implication.} Negative results at the
representation-structure level of latent reasoning are useful
because they rule out architectural families cheaply. The
structural argument in \S\ref{sec:flat} and the positive control in
\S\ref{sec:pc} together close off readout-architecture fixes for
matrix-CODI's rank-blindness. A designer who wants matrix latents to
\emph{use} their rank as a functional observable should plan on
changing the training objective, not the readout.
