\section*{Reproducibility}
\label{sec:repro}

All training, evaluation, and probe code is released anonymously at
\url{https://anonymous.4open.science/r/matrix-codi-rank-blindness}.
The release includes:

\begin{itemize}
\setlength\itemsep{2pt}
\item \texttt{run\_matrix\_codi.py}: the matrix-CODI training
      script. The \texttt{MatrixBottleneck} class implements the
      $W_{\text{up}}\!\to\!\text{reshape}\!\to\!\text{thinker}\!\to\!\text{flatten}\!\to\!W_{\text{down}}$
      pipeline. All five readouts in \S\ref{sec:pc} are selectable
      via the \texttt{--readout} flag.
\item \texttt{probe\_z.py}: the linear probe pipeline for
      \S\ref{sec:flat}. Produces the AUC numbers in
      Fig.~\ref{fig:probe}.
\item \texttt{rank\_eval.py}: the rank-$k$ projection evaluator.
      Given a checkpoint, it computes the accuracy-vs-$k$ curve and
      the Spearman correlation between per-sample effective rank
      and correctness.
\item Raw rank-$k$ evaluation output files (JSON) for the four
      positive-control readouts, matching Table~\ref{tab:pc} and
      Fig.~\ref{fig:pc}.
\item A human-readable experiment log with per-run hyperparameters,
      rank trajectories, and wall-clock times.
\end{itemize}

Datasets: ProsQA \citep{hao2024coconut} from the
\texttt{facebookresearch/coconut} release. GSM8K-Aug
\citep{shen2025codi} from the CODI release.

Backbone: pretrained GPT-2 small (124M), medium (355M), and large
(774M) from the standard public checkpoints.

Training hardware: a single NVIDIA H100 (80GB HBM3) per run. All
reported numerical results in the main body trace to a specific
checkpoint and evaluator run in the release.

\paragraph{Headline numbers by source.} Table~\ref{tab:four-flat}
rows R1--R3b from the matrix-CODI training script
\texttt{run\_matrix\_codi.py} with seed $1337$;
Fig.~\ref{fig:seed-decoupling} from the same script with seeds
$\{1337,42,7\}$; Table~\ref{tab:pc} and Fig.~\ref{fig:pc} from
\texttt{run\_matrix\_codi.py} with \texttt{--readout} in
$\{$\texttt{flatten},\texttt{bilinear},\texttt{bilinear\_gelu},\texttt{svd\_aug},\texttt{quadratic}$\}$
evaluated via \texttt{rank\_eval.py}; Fig.~\ref{fig:probe} from
\texttt{probe\_z.py} on the Round 3 $\gamma=0$ checkpoint;
Figs.~\ref{fig:depth} and~\ref{fig:scale} from the vanilla
SFT and vanilla CODI scripts in the same release.
