\section{The Bound and the Teeth That Keep It a Bound}
\label{sec:setup}

\textbf{Task and bound.} Each episode presents $K$ key$\to$value
bindings, then queries one; keys and values are fresh per episode
(Gaussian, normalized, or exactly orthonormal following
\citealp{mezzadri2007random}), so evaluation uses never-seen bindings. Exact recovery $Z k_j = v_j$
for $K$ bindings with linearly independent keys and values forces
$Z \Kmat = \Vmat$, hence the classical bound
$K = \rank(\Vmat) = \rank(Z \Kmat) \leq \rank(Z)$
\citep{kohonen1972correlation,anderson1972simple}. We measure whether a
trained, bottlenecked model \emph{discovers} this rank under plain SGD,
and whether it is causally load-bearing.

\textbf{Tooth 1: exact continuous recovery, never argmax.}
\citet{nichani2025factual} construct a rank-$m$ linear memory storing
approximately $md$ associations under interference-tolerant argmax
decoding; under that rule a rank-1 state passes at every $K$ we test
and the bound collapses. We therefore pin the decoder to the linear
unbind itself, $\mathrm{pred} = Z \cdot k_j$, scored by an absolute bar
($\recninety$: the fraction of queries with $\cos > 0.9$), never an
argmax and never a learned readout that could launder rank.

\textbf{Tooth 2: a single-state bottleneck, verified behaviorally.}
``Hold $K$ items'' is satisfiable in full attention by attending to
$K$ raw-input positions at rank~1 each, so all information
funnels through one $d \times d$ state and the decoder is a pure
function of $(Z, \mathrm{query})$: after the encoder writes $Z$, the
raw-input cache is corrupted and the decode is confirmed bit-for-bit
unchanged (a gradient blank-out test, not a mask-shape check).
% <!-- evidence: R10 -->

\textbf{Model and controls.} A Transformer encoder over the binding
tokens, with $d$ learned row-reader latents producing the rows of
$Z \in \R^{d \times d}$: permutation-invariant, able to express any
rank up to $d$, and not hard-coding $\sum_j v_j k_j^\top$. Every model
is under 1M parameters (the $d = 16$ composition model: 170{,}896).
% <!-- evidence: R9 -->
Effective rank (entropy of the normalized singular-value spectrum) is
the pre-registered primary metric; the primary causal control
constrains $Z$ to rank $k$ at \emph{every training step} (spectral
projection), not post hoc.
