\section{Related Work, Limitations, and Outlook}
\label{sec:related}

\citet{nichani2025factual} is the nearest neighbor: a hand-built
existence construction under interference-tolerant argmax decoding;
this work pins the readout so that construction cannot pass, then
measures and intervenes on the rank SGD recruits
(\citealp{barnfield2026sharp} give static-memory thresholds).
\citet{nazari2026rank} and \citet{sun2026staterank} measure
effective-rank dynamics in pretrained linear-attention models
observationally (an \emph{upper} bound $\rank(S_t) \leq t$); here the
write count is controlled and necessity is causal. \citet{mishra2026m2rnn} train matrix-state RNNs on $S_3$ with no rank
measurement or intervention (their own vector-state control matches);
\citet{grazzi2025negative} and \citet{siems2025deltaproduct} (the
Householder extension of \citealp{yang2024deltanet}) characterize the
per-token \emph{transition} operator, not the trained \emph{state's}
own rank, the quantity here. Continuous-CoT capacity
results are existence constructions, not measurements
\citep{zhu2025superposition,gozeten2025cot2}; the
compositional-generalization caution
\citep{dziri2023faithfate,wang2024grokked} is met with an exact
eigenstructure match.

\textbf{Limitations.} Every experiment is synthetic and under 1M
parameters; generality at scale, and across state-writing architectures
(recurrent fast-weight, attention-readout, associative-memory), is
untested.
% <!-- evidence: R9 -->
The depth-amplification contrast rests on the single converged
force-rank-$(K{-}1)$ seed; the other two at that cell collapsed under
a documented numerical instability in the spectral-projection backward
pass, so it is a case study consistent with the spectral prediction,
not an $n{>}1$ mechanism; the $d \geq 48$ frontier cells rest on
single seeds. The registered falsifiers stand
(Appendix~\ref{app:repro}). A concurrent companion submission
(group-composition state tracking) shares no figures or tables but
restates this paper's recruitment, causal-necessity, and composition
numbers as background for its distinct contribution, a group-dimension
rank law this paper does not make.
