\appendix

\section{Depth 21 Under the Single $K$-Cycle}
\label{app:period}

Under a single 8-cycle, $\pi^{21} = \pi^{21 \bmod 8} = \pi^{5}$, so the
nominal depth-21 probe shares its \emph{target} with the already-tested
depth 5; the two are not distinct group-theoretic queries. What depth 21
measures is numerical stability under 21 sequential self-applications of
the trained operator. In the four converged seeds $\recninety$ reads
$\geq 0.9996$ at both $h{=}5$ and $h{=}21$, so repeated self-application
does not amplify their residual error;
% <!-- evidence: R3 -->
in the rank-starved operator the same 16 additional applications
compound a passing shallow-hop cosine into collapse (0.947 at $h{=}5$
to 0.060 at $h{=}21$ in recovered fraction),
% <!-- evidence: R4 -->
which is exactly the amplification claim of \S\ref{sec:compose}, not a
generalization claim to unseen permutation targets.

\begin{table}[!htb]
\centering
\small
\begin{tabular}{cccc}
\toprule
$h$ & predicted $\cos$ & measured $\overline{\cos}$ & measured $\recninety$ \\
\midrule
1  & 0.9317 & 0.9303 & 0.996 \\
3  & 0.9258 & 0.9259 & 0.986 \\
5  & 0.9181 & 0.9212 & 0.947 \\
7  & 0.9089 & 0.9163 & 0.881 \\
21 & 0.7536 & 0.8206 & 0.060 \\
\bottomrule
\end{tabular}
\caption{The rank-starved ($k = 7 = K - 1$) operator's depth-decay
curve: cosine predicted from the trained operator's own eigenspectrum
(no raw keys) against the measured values, with the recovered fraction
alongside. Every hop through $h = 7$ matches within 0.008; at $h = 21$
the per-item distribution widens and the exact-match metric collapses
while mean cosine moves little.}
% <!-- evidence: R4 -->
\label{tab:depthcurve}
\end{table}

\section{Per-Seed Subspace Decomposition}
\label{app:subspace}

\begin{table}[!htb]
\centering
\small
\begin{tabular}{lccccc}
\toprule
seed & $\mathrm{effrank}(A)$ & residual & leakage (\% of $\|Z\|$) & $\mathrm{effrank}(D)$ & $\rho(D)$ \\
\midrule
s1 & 8.000 & 0.7\% & 0.44\% & 8.000 & 1.38 \\
s2 & 7.999 & 1.9\% & 1.06\% & 8.000 & 1.27 \\
s3 & 7.999 & 1.5\% & 0.87\% & 8.000 & 2.86 \\
s4 & 7.999 & 2.4\% & 1.45\% & 7.999 & 1.02 \\
\bottomrule
\end{tabular}
\caption{Entity-subspace block decomposition of the four converged
$K{=}8$ composition seeds (means over four evaluation episodes),
recomputed from the archived $Z$-dumps. $A$ is the operator restricted
to the $K$-dimensional entity subspace; residual is the scale-corrected
Frobenius distance to the ideal cycle after fitting the single isotropic
scale that cosine scoring cannot see; leakage is the cross-block norm
$\|C\|$ as a fraction of $\|Z\|$; $D$ is the complement block, full-rank
and non-contractive ($\rho(D) \geq 1$) yet never visited by a query
trajectory that starts in the entity subspace.}
% <!-- evidence: R5 -->
\label{tab:subspace}
\end{table}

\section{Reproducibility}
\label{app:repro}

Training, evaluation, and analysis code is available at
\url{https://anonymous.4open.science/} (anonymized for review). Raw
per-run JSON archives (dense training-step checkpoints, not summary
rollups) back every number in this abstract; each figure regenerates
from those archives through a single versioned script that asserts an
md5 checksum for every source file it loads, so a changed artifact fails
the build rather than silently plotting stale data. Hypotheses, metrics,
force-rank controls, and decision bands were registered in design
documents before training, and every deviation (one metric
re-registration, logged before the affected sweep) is recorded inline
in those documents. The registered falsifiers stand open:
force-rank-$(K{-}1)$ at $\geq 0.9\times$ ceiling at multiple $(d, K)$
points, or a $d \geq 32$ cell clearing the 0.7 bar, would overturn the
causal claim.
The force-rank staircase aggregate retains one recovery figure per
$(d, K, k)$ cell, not per-seed instability diagnostics; the necessity
direction (recovery $\approx 0$ below $K$) is robust to the cause of
any individual sub-$K$ collapse, while the smoothness of the post-knee
approach should be read at the aggregate level only.
The analysis pipeline emits benign BLAS \texttt{RuntimeWarning}s from
near-singular intermediate products on some platforms; every
decomposition output is asserted finite before use, so these do not
affect any reported number.
