
\textbf{Limitations.} All results are on sub-1M-parameter synthetic
testbeds, not claims about pretrained language models. Razor cells are
single-seed except $S_3$ (four seeds); $S_4$/$S_5$'s anchor-bar margins
and $S_3$/$S_5$'s soft convergence at the shortest pinned budget
(Appendices~\ref{app:m1}, \ref{app:gate1a}) both read directionally, not
as an unqualified 5/5.

\appendix

\section{The Five Groups and Reference Representations}
\label{app:groups}

\begin{table}[h]
\centering
\small
\begin{tabular}{lcccl}
\toprule
$G$ & $|G|$ & $\dmin$ & solvable & realization of $\rho_G$ \\
\midrule
$S_3$ & 6   & 2 & yes & standard (dihedral) \\
$S_4$ & 24  & 3 & yes & cube rotations \\
$A_5$ & 60  & 3 & no  & icosahedral rotations \\
$S_5$ & 120 & 4 & no  & standard (zero-sum) \\
$A_6$ & 360 & 5 & no  & standard (zero-sum) \\
\bottomrule
\end{tabular}
\caption{The five task groups. $\dmin$ is the minimal faithful real
representation dimension; each reference representation is orthogonal by
construction and verified faithful (injective on $G$) at build time.
$S_4$/$A_5$ form the marquee pair: matched $\dmin$, opposite solvability.}
\label{tab:groups}
\end{table}

\section{Per-Seed Observational Rank}
\label{app:m1}

\begin{table}[h]
\centering
\small
\begin{tabular}{lccccc}
\toprule
$G$ & $\dmin$ & $n$ & restricted eff.\ rank & band & in band \\
\midrule
$S_3$ & 2 & 3 & $1.877 \pm 0.060$ & $[1.4, 2.6]$ & 3/3 \\
$S_4$ & 3 & 5 & $2.852 \pm 0.054$ & $[2.1, 3.9]$ & 5/5 \\
$A_5$ & 3 & 5 & $2.832 \pm 0.062$ & $[2.1, 3.9]$ & 5/5 \\
$S_5$ & 4 & 3 & $3.591 \pm 0.069$ & $[2.8, 5.2]$ & 3/3 \\
$A_6$ & 5 & 3 & $4.736 \pm 0.023$ & $[3.5, 6.5]$ & 3/3 \\
\bottomrule
\end{tabular}
\caption{Restricted effective rank of the unconstrained trained state
(mean $\pm$ sd over seeds) against each group's $\dmin$, with the
pre-registered $[0.7, 1.3] \cdot \dmin$ band. Every seed of every group is
in band; Spearman $\rho = 0.9747$ is the tie-capped maximum; the
dimension-matched pair $S_4$/$A_5$ differs by 0.019 rank-units.}
% <!-- evidence: N4 -->
\label{tab:m1}
\end{table}

As a fraction of the geometrically forced ceiling
$\sqrt{(\dmin{-}1)/\dmin}$, the below-$\dmin$ crosscheck cosines read
$S_3$ 86.3\%, $S_4$ 91.2\%, $A_5$ 94.9\%, $S_5$ 75.7\%, and $A_6$
93.5\% (mean 88.3\%). $S_5$ is the outlier, consistent with its
soft-convergence flag (Appendix~\ref{app:gate1a}).
% <!-- evidence: N12 -->

Per-group deviation of the mean from $\dmin$ (Table~\ref{tab:m1}):
$S_3$ 6.1\%, $S_4$ 4.9\%, $A_5$ 5.6\%, $S_5$ 10.2\%, $A_6$ 5.3\%.
% <!-- evidence: N4 -->

$S_4$ and $S_5$'s razor $k{=}\dmin$ cells (Figure~\ref{fig:tracking},
main text) numerically exceed their own unconstrained anchor by 0.10--0.15;
plausibly the capped arm skips the ambient dimensions the zero-padded
anchor must still learn to null under the same fixed step budget, an
unconfirmed hypothesis.
% <!-- evidence: N14 -->

The instrument's centering step is load-bearing: an uncentered lens
scores a flawless synthetic model 0.705 where the centered production
lens scores 0.9996 on identical data.
% <!-- evidence: N11 -->
The disclosed length-robustness split referenced in
Section~\ref{sec:observed} scores only words of length $L \geq 2$ (the
$L = 1$ read is attention-degenerate and was demoted with a mechanism
note in the run records): it moves per-group mean recovery cosine by at
most 0.013 (per-seed at most 0.041), with no group-selective divergence.
% <!-- evidence: N10 -->
Per-seed, $S_3$'s $k{=}\dmin$ cell clears its own seed's $0.9\times$anchor
bar in 2 of 4 seeds (seed 1: $+0.010$; seed 3: $+0.110$; seed 0: $-0.045$;
seed 2: $-0.120$); the self-referential bar ($0.9\times$ the seed-mean
anchor, 0.574) narrowly fails the seed-mean $k{=}\dmin$ value (0.5625),
which is why the main text uses the fixed pre-registered literal (0.495)
rather than a self-referential recompute.
% <!-- evidence: N3 -->

\section{The Ambient-Identity Tax in Detail}
\label{app:damb}

The identity-padded force-rank target
$\rho_G(\cdot) \oplus I_2$ has rank $\dstate = \dmin + 2$ for every
group. Because the identity block is constant across words, it is the
cheapest reliable loss reduction available, and a rank-capped arm buys it
before the group representation; the residual budget for $\rho_G$ is
$k - 2$, below $\dmin$ at every grid point, so no capped cell under
identity-padding ever tested the law's confirm direction. Three raw-artifact
signatures established this: (i) force-rank cells' direct cosine matched
the rank-$k$ optimum $\sqrt{k/\dstate}$ with mean absolute deviation 0.028
across all 39 cells (maximum 0.166; the two $S_5$ below-$\dmin$ cells are
the only outliers, an additional disclosed optimization shortfall), so the
arms trained to their rank-constrained ceiling rather than under-training;
% <!-- evidence: N7 -->
(ii) centered restricted rank of capped arms landed near $k - 2$, the
residual budget; (iii) in the zero-padded grid, a deliberately
eye-padded corroboration arm reproduces the failure on demand at raw
$k = \dmin{+}1$ (effective budget $\dmin{-}1$) while the tax-paid point
recovers.
% <!-- evidence: N1 -->
The zero-padded grid's 30 cells were configuration-verified one-by-one
against a manifest re-derived independently from the design record (steps,
padding mode, and force-rank value per cell; zero skipped steps).
% <!-- evidence: N8 -->

\section{Soft Convergence at the Shortest Pinned Budgets}
\label{app:gate1a}

\begin{table}[h]
\centering
\small
\begin{tabular}{lccccc}
\toprule
$G$ & anchor & $k{=}\dmin{-}1$ & $k{=}\dmin$ & $k{=}\dmin{+}1$ & bar (0.92) \\
\midrule
$S_3$ & 0.914 & 0.665 & 0.900 & 0.903 & fail (all 4) \\
$S_5$ & 0.876 & 0.801 & 0.879 & 0.877 & fail (all 4) \\
\bottomrule
\end{tabular}
\caption{Convergence health (gate1a, min validation cosine over
$L \in [2, 5]$) for every $S_3$/$S_5$ razor cell at the shortest pinned
step budget. All eight cells fall short of the 0.92 margin; $S_5$'s
0.876--0.879 range is a larger shortfall than $S_3$'s 0.900--0.914. The
necessity leg (Section~\ref{sec:razor}) is unaffected by construction; the
sufficiency leg for these two groups should be read as directionally
consistent under disclosed soft convergence.}
% <!-- evidence: N13 -->
\label{tab:gate1a}
\end{table}

\section{Depth-21 Periodicity Under the Single $K$-Cycle}
\label{app:period}

The binding task's held-out composition probe (Section~\ref{sec:setup})
uses a single Hamiltonian $K$-cycle, chosen because a general permutation
decomposes into short disjoint cycles that collapse held-out hops into
trivial or in-distribution queries. Under a single 8-cycle, $\pi^{21} =
\pi^{21 \bmod 8} = \pi^5$ exactly, so nominal depth 21 shares its target
with the already-tested depth 5; the two are not distinct group-theoretic
targets. What depth 21 measures instead is numerical stability under 21
sequential self-applications of the trained operator: in the four
converged seeds, $\recninety$ at $h=5$ and $h=21$ both read $\approx 1.0$,
a non-trivial statement that repeated self-application does not amplify
error; the fifth (still-transitioning) seed reads 0.303 at $h=5$ and
0.0001 at $h=21$, the same underlying imperfection compounding under
repetition.
% <!-- evidence: N9 -->
