\begin{abstract}
Native Composition Reads (NCR) writes relation operators into a $d\times d$
fast-weight state from context and reads a depth-$h$ composition via
$O(\log h)$ repeated squaring instead of an $O(h)$ sequential rollout --
the load-bearing claim is query-time access complexity, not a new
state-tracking expressivity result (matrix-valued recurrent state already
provably escapes $\mathrm{TC}^0$ elsewhere in the published record). At
$K{=}8$ entities, NCR holds exact recovery to the pre-registered separation
depth $h^\ast{=}61$ (median $1.0$) while the strongest $O(h)$ baseline has
already decayed to $0.158$, a full band separation on every claim-eligible
residue. % evidence: NOVEL_ARCH_WATERFALL.md S7e
Scaling the same recipe to $K{=}32$ closes: sixteen dimension/budget cells
and eight further budget-rescue cells all fail to reach robust
convergence, and far-depth recovery reads exactly $0.0000$ in every one.
% evidence: NOVEL_ARCH_WATERFALL.md S11.5/S11.6
We diagnose the wall rather than accept it as fundamental: the write is
trained with a cosine loss at $h\le 3$ that constrains only direction, not
conditioning, so gradient descent has no pressure to keep the written
operator normal. The trained $K{=}32$ operator is measurably non-normal
(departure-from-normality $0.055$--$0.063$) and ill-conditioned (condition
number $321$--$2952$ across the tested $d$-arms, against $\approx 1$ for a
healthy $K{=}14$ operator), and the weakest eigenmode is annihilated by
$h\approx 6$. % evidence: NCR_ORTHO_WRITE.md S1
Projecting the already-trained operator to its nearest orthogonal matrix,
with no retraining, extends the surviving depth from $h\approx 6$ to
$h\approx 27$--$51$ and lifts mid-depth recovery from near-zero to
$0.73$--$0.98$, showing the wall is a synthetic property of the objective,
not the architecture's capacity (effective rank already sits at $\approx K$
in every dead cell). % evidence: NCR_ORTHO_WRITE.md S5
We pre-registered WIN/PARTIAL/NULL/FAIL bands for an experiment training
that constraint directly, before any result existed; the verdict landed
FAIL -- the Newton--Schulz-constrained write does not train (in-distribution
recovery $0.000$ in $4/4$ seeds at both $K$ tested, the pre-registered
``too rigid to train through'' mechanism) while the free-write control
trains cleanly in the same harness -- conditional on an in-progress code
re-audit that has not yet excluded an implementation defect in the
projection. A failed pre-registered rescue under a validated instrument is
itself evidence about the wall's character. Every model in this
paper is under 200K parameters; this is a toy-scale mechanistic study, and
we make no scaling claim beyond the diagnosis itself.
\end{abstract}
