\section{The Fix Experiment: Verdict}
\label{sec:pending}

The diagnosis in Section~\ref{sec:diagnosis} is a post-hoc counterfactual
on an already-trained, free-write operator -- it does not by itself
establish that a model can learn to write an orthogonal operator
directly. A GPU wave training the Newton--Schulz-constrained write
end-to-end at $K \in \{24, 32\}$, $d{=}K{+}1$, 4 seeds, $4\times$ budget,
plus a structured-operator discriminator (composing paths of
\emph{distinct} random-orthogonal operators, removing every periodicity
handle so depth cannot collapse modulo a cycle length) was pre-registered
before any tunable code existed and has now run to completion and been
read.
% evidence: NCR_ORTHO_WRITE.md S2/S4b/S8/S9

\begin{quote}
\textbf{VERDICT (conditional on an in-progress code re-audit -- see
below): Part A FAIL. Part B FAIL.} The Newton--Schulz-constrained write
does not train at $K{=}32$; the free-write control, run in the identical
harness, reproduces the wall cleanly, validating the instrument.
\end{quote}
% evidence: NCR_ORTHO_WRITE.md S9.5

The bands below were fixed before any result existed and are reproduced
verbatim so the verdict below can be checked mechanically against them; a
FAIL is a finding about the constraint, not a result to hide.
% evidence: NCR_ORTHO_WRITE.md S6 (record-before-read discipline)

\textbf{Part A (primary, $K{=}32$, realistic re-registered $h^\ast{=}40$
-- $13\times$ the training depth, chosen because the diagnosis's polar
frontier at $4\times$ budget already reaches $h\approx 51$).}
\textbf{WIN}: ortho-write median $\mathrm{recovered\_frac@0.9}$ at
$h^\ast{=}40 \geq 0.9$ across converged seeds, AND the pinned free-write
baseline reads $< 0.5$ at $h{=}40$ (already established: the free-write
arm is DEAD at every depth $\geq 29$, Section~\ref{sec:wall}); the
mechanistic signature must corroborate (departure-from-normality
$\leq 0.02$, condition number $\lesssim 2$, $\min|\lambda|/c^\ast \geq
0.9$). \textbf{PARTIAL}: median $\geq 0.9$ at $h \in \{20, 29\}$ but $<0.9$
at $h{=}40$ -- a genuine $\geq 3\times$ extension of the free-write
frontier ($\approx 6$), short of the full target. \textbf{NULL}: median
$< 0.9$ at every depth $\geq 12$ while the write still trains (Gate-0
passes) -- the constraint trains but buys no far-depth gain; the mechanism
would be wrong or incomplete. \textbf{FAIL}: Gate-0 DEAD in $\geq 3/4$
seeds -- the constraint is too rigid to train through.
% evidence: NCR_ORTHO_WRITE.md S3/S4

\textbf{Part B (structured-operator discriminator, $K{=}32$, $R{=}4$
distinct operators, $L^\ast{=}32$ path compositions, mod-$K$-trap-safe by
construction: distinct operators per hop, no consecutive repeats,
fixed-point exclusion).} Identical WIN/PARTIAL/NULL/FAIL bands with path
length $L$ in place of $h$. Prediction, stated before the run: a product
of $L$ orthogonal matrices is exactly orthogonal for every $L$, so this
discriminator should show the sharpest possible separation if the
mechanism is real, with no residual eigenphase question the way a single
cycle's periodicity leaves open.
% evidence: NCR_ORTHO_WRITE.md S4/S4b

\textbf{Pre-registered odds (reproduced for the record; not a prediction
dressed as a result -- the actual outcome is reported below).}
$\approx 70$--$75\%$ Part A reaches WIN or high PARTIAL, on the basis that
the mechanism is measured (not hypothesized) and the in-silico polar
preview already achieves comparable depth extension without retraining;
residual risk is the constraint blocking Gate-0 convergence, or trained
orthogonality not being tight enough over $40$ physical applications.
$\approx 80\%$ Part B reaches WIN, higher than Part A because exact
orthogonality of a product carries no residual-phase caveat. $\approx
20$--$25\%$ the original synthetic $h{=}253$ target is recovered --
explicitly not the target, reported only if it happens. The residual risk
named for Part A -- Gate-0 blocking convergence -- is exactly what
occurred. % evidence: NCR_ORTHO_WRITE.md S7

\subsection*{Result}

\textbf{Part A: FAIL.} Gate-0 (in-distribution trainability) is DEAD in
$4/4$ seeds at $K{=}32$: $\mathrm{recovered\_frac@0.9} = 0.000$ at
$h \in \{1,2,3\}$ in every seed (bar $\geq 0.9$; also below the $0.5$
FAIL threshold), and mean $A_{\mathrm{eff\_rank}}$ ($17.6$--$27.3$) never
reaches the $28.8$ rank bar. This is the \textbf{FAIL} band by
construction: Gate-0 DEAD in $\geq 3/4$ seeds. WIN, PARTIAL, and NULL are
all mechanically unreachable at this reading -- NULL specifically
requires Gate-0 to PASS, which it does not. The secondary, not-separately-
scored $K{=}24$ cells fail identically (4/4 Gate-0 dead, same signature).
% evidence: NCR_ORTHO_WRITE.md S9.1

\textbf{The training signature matches the pre-registered mechanism, not
a dead-on-arrival run.} Every ortho cell shows the same shape: loss dips
transiently (minimum $0.063$--$0.909$ across cells) and then collapses
back to $\approx 1.0$ (random) by the end of training. The optimizer
briefly engages the task before the orthogonality constraint destabilizes
it -- the ``too rigid to train through'' failure mode named in \S4 in
advance, observed here rather than assumed.
% evidence: NCR_ORTHO_WRITE.md S9.1

\textbf{The free-write arm, run in the identical harness, validates the
instrument.} At $K{=}24$ it recovers essentially perfectly at every
tested depth ($\approx 1.0$ from $h{=}5$ to $h{=}61$, spectrally healthy:
departure-from-normality $\approx 0.006$, condition number $\approx 1.0$).
At $K{=}32$ it reproduces the wall Section~\ref{sec:wall} and
Section~\ref{sec:diagnosis} already describe: in-distribution recovery
$0.875$--$0.991$ at $h{=}1,2,3$, then $\mathrm{recovered\_frac@0.9} =
0.000$ at every $h \geq 20$, with the same non-normal, ill-conditioned
spectral signature (departure-from-normality $0.047$--$0.060$, condition
number $54.9$--$558.6$) as the archived checkpoints
Section~\ref{sec:diagnosis} analyzed. Because free-write trains cleanly
and ortho-write does not, in the same pipeline, on the same task, the
failure localizes to the Newton--Schulz-constrained write path, not to
the harness, the eval, or the task itself.
% evidence: NCR_ORTHO_WRITE.md S9.1/S9.3

\textbf{Part B: FAIL, and uninformative by construction.} The structured-
operator discriminator's ortho-bank arm is Gate-0 DEAD in $4/4$ seeds
($\mathrm{recovered\_frac@0.9} = 0.000$ at $L \in \{1,2,3\}$, loss flat at
$\approx 1.0$ throughout) -- FAIL per \S4b's identical bands. Separately,
and disclosed as its own finding rather than folded into the Part-A
result: the \emph{free}-bank baseline also failed Gate-0 in $4/4$ seeds
($3/4$ seeds never leave loss $\approx 1.0$; the fourth reaches
in-distribution $0.412$ at $L{=}1$ before decaying, still below the $0.5$
bar). Neither arm learned even the in-distribution ($L{=}1,2,3$) task at
this configuration ($R{=}4$ operator bank, 320K steps), so Part B does
not produce the ortho-vs-free far-depth contrast it was designed to
discriminate. This is a bank-task calibration issue at this
configuration, disclosed as such -- it does not strengthen or weaken the
Part-A ortho-write FAIL.
% evidence: NCR_ORTHO_WRITE.md S9.2

\textbf{Caveat, stated plainly: this verdict is conditional on an
in-progress code re-audit.} The behavioral signature -- loss engaging
then collapsing, free-write training cleanly in the identical pipeline --
is consistent with the pre-registered ``too rigid to train through''
mechanism, and that is the reading this section reports. But behavioral
data alone cannot exclude an implementation defect in the 40-iteration
Newton--Schulz polar projection: a CPU smoke test passed (step-0 gradient
norm $0.5$--$0.9$) yet training still destabilized over the full
320K-step trajectory, and an independent re-audit of the
Newton--Schulz/\texttt{encode} path under the full optimizer trajectory
is underway but not complete as of this draft. The FAIL verdict above is
behaviorally the pre-registered mechanism; it is not yet ruled
independent of a possible implementation bug, and the paper's verdict is
conditional on that audit closing clean. This caveat is recorded in
\texttt{NCR\_ORTHO\_WRITE.md} \S9.4 (item 1) and must be read alongside
the verdict, not as a footnote to it.
% evidence: NCR_ORTHO_WRITE.md S9.4

\textbf{What this establishes.} The $K{=}32$ wall reproduces under an
independent training run in this section's own harness and is not
rescued by the pre-registered orthogonal-write fix at this
parametrization. A failed rescue attempt, run under a validated
instrument (the free-write control trains cleanly in the same pipeline),
is itself evidence about the wall's character: it is not merely that
nobody has tried constraining the write, but that the specific,
principled constraint this program pre-registered does not survive
contact with training at $K{=}32$, pending the code audit above. Per the
pre-registered next move on FAIL, the follow-on is a fallback
parametrization (a Cayley map or skew-symmetric matrix exponential,
Section~\ref{sec:diagnosis}) or a softer spectral-penalty variant -- not
more training budget, which Section~\ref{sec:wall} already showed does
not move this wall.
% evidence: NCR_ORTHO_WRITE.md S9.5
