\section{The Honest Open Question}
\label{sec:open-question}

\begin{itemize}
  \item Neither of the two most naturally testable single-variable accounts
    of the cliff's dissolution survives: the $K/\dstate$ ratio (Act 2)
    is ruled out directly by the flat $d=128$ result, and scalar,
    rank-4-structured anchor-table coherence (Act 3) is ruled out directly
    by the flat dose-response result --- injecting coherence up to and
    beyond $d=64$'s own realized training band still does not reproduce
    the cliff at $d=128$.
  \item \textbf{Two candidates survive the waves above; a follow-up wave
    (below) has since tested one of them further.}
  \begin{enumerate}
    \item \textbf{Overlap structure (diffuse vs.\ concentrated).} A
      concentrated (rank-4) injection collapses the table's $n$ rows toward
      a shared low-dimensional subspace; a \emph{diffuse}
      (\texttt{subspace\_rank=48}) injection spreads the same scalar dose
      across a far less structured set of directions. It remains possible
      the cliff tracks \emph{how} coherence is distributed across the
      table, not merely its scalar magnitude. This is a co-primary,
      pre-registered follow-on arm --- design complete, not yet run.
    \item \textbf{Absolute state capacity.} $d_{\mathrm{state}}$ grew
      $4\times$ (64$\to$128) between the cliff's location and its
      disappearance, while $K$ only grew $\sim\!2\times$ (34--46 at $d=64$
      vs.\ 68--92 at $d=128$, holding $K/\dstate$ fixed). It remains
      possible the relevant quantity is raw available state dimensionality
      itself (or slack $\dstate - K$), not any ratio or coherence measure
      --- a hypothesis no wave here can speak to, since $d_{\mathrm{state}}$
      was never varied independently of $K/\dstate$ or coherence.
  \end{enumerate}
  \item \textbf{Candidate 2 has since been tested directly, and the
    ratio-invariance null it would need is REFUTED --- load tolerance
    grows SUPER-LINEARLY in $d_{\mathrm{state}}$, not at a fixed
    $K/\dstate$ ratio.} A follow-up wave re-ran the identical localization
    fit at two further scales. Against a pre-registered ratio-invariant
    band $[0.4745, 0.6165]$ (centered on $d{=}64$'s own $x_0{=}0.5455$),
    $d_{\mathrm{state}}{=}80$ locates its own transition at
    $x_0{=}0.6779$ (95\% CI $[0.6683, 0.6867]$, a clean, non-degenerate
    fit, $n{=}5$ seeds/$K$) --- entirely above the band, a real exclusion
    of fixed-ratio invariance. \textbf{At $d_{\mathrm{state}}{=}96$, a
    wide-grid attempt ($K{=}72,78,84,90$) initially flagged 11 of 12 new
    cells inadmissible; an independently-audited repro instrument
    diagnosed a tolerance miscalibration in the convergence gate --- all
    4,608 flagged episodes resolve by $n_{\mathrm{iter}}{\leq}28$ against a
    production $n_{\mathrm{iter}}{=}20$ --- and only the admission flag,
    never the $h4$ measurement, was recalibrated, unlocking the full grid
    at zero new GPU spend.} The re-fit on the complete grid
    ($K{\in}\{69,72,78,84,90\}$, $n{=}3$/$K$) finds NO cliff anywhere
    through $K/\dstate{=}0.9375$, but a non-monotonic curve ($h4$ means
    $0.9592/0.9216/0.9326/0.9581/1.0000$: a dip at $K{=}72/78$, partial
    recovery at $K{=}84$, an exact ceiling at $K{=}90$) that the sigmoid
    fit cannot localize --- 100\% of 4,000 bootstrap resamples land at the
    fit's own degenerate parameter bounds, from genuine per-seed scatter
    (the noisiest group, $K{=}72$, spans $0.8426$--$0.9904$) rather than a
    data gap. An absolute-slack account ($x_0 \propto d_{\mathrm{state}}$
    minus a constant, band $[0.718,0.739]$) and a power-law account (band
    $[0.768,0.837]$) both remain live and neither is excluded ---
    \textbf{the precise functional form of the super-linear growth, and
    the exact $d_{\mathrm{state}}{=}96$ transition point, remain open.}
    Seed escalation at $K{=}72$ (the noisiest group) is the registered
    next step, not launched by this wave.
  \item Adjudicating candidate 1 still requires a not-yet-designed wave: a
    diffuse-structure Stage 2 (co-primary with the rank-4 result already
    reported). Not registered in detail as of this writing; a natural,
    concrete next step, stated here rather than left implicit.
  \item We report this as a genuinely open question, not a hedge: a
    precisely-located transition, its clean dissolution with scale, and
    the ruling-out of the single most natural mechanistic account of that
    dissolution is, on its own, a complete and informative capacity story
    --- even though it does not yet identify what \emph{does} drive the
    transition.
\end{itemize}
