\section{Related Work and Limitations}
\label{sec:related}

\looseness=-1 \textbf{Related work.} \citet{silberzahn2018many} and
\citet{damour2020underspecification} document divergent conclusions
from underspecified pipelines; here the models are fixed and
the \emph{measurement pipeline} is the adjudicated object.
\citet{henderson2018deep} and
\citet{bouthillier2021accounting} quantify seed and tuning variance;
none of the six defects is sampling variance, so variance
accounting catches none. REFORMS
\citep{kapoor2023reforms}, \citet{sculley2018winners}, and
\citet{lipton2018troubling} prescribe what a trustworthy paper
discloses; we supply the in-flight machinery such checklists
presume exists. The probing literature supplies the
instrument-hygiene checks the wrong-layer case rediscovers at
evaluation scale: control tasks and
selectivity~\citep{hewitt2019control}, the probe-versus-behavior
dissociation~\citep{belinkov2022probing}, amnesic
probing~\citep{elazar2021amnesic}, and randomization-based sanity
checks~\citep{adebayo2018sanity}. \citet{schaeffer2023emergent}
argue the inverse of our thesis (an apparent model \emph{property},
emergence, as metric artifact) from the same
instrument-first stance; each diagnoses one instrument
class, where this catalogue is a serial, pre-registered,
cross-instrument adjudication record with falsifier teeth.
\citet{nichani2025factual} prove Rule 5's argmax/exact-recovery gap;
we report it firing live (\S\ref{sec:wronglayer}); the fast-weight
testbeds follow \citet{schlag2021linear}.

\looseness=-1 \textbf{Limitations.} All six incidents come from one program, one
team, one architecture family: they support no prevalence claim, and
we cannot report the rate a lighter process would miss.
The catalogue is survivor-biased: a defect missed by both crosscheck
and falsifier battery appears in no record we could write. It is
biased by trigger too: the discipline fires on apparent model
failure, so defects flattering the model are under-represented; five
of six fixes moved results in the program-favorable direction. In the
one incident costed, the two diagnostic instruments took $0.50$ and
$1.10$ GPU-hours (archived wall-clock) against the $6.33$ GPU-hours
of training the uncorrected readings would have discredited. We
report no false-alarm rate: the records do not tally escalations
against clean instrument-health passes, so the precision of
``suspect the instrument first'' is unmeasured.
% <!-- evidence: I6 -->
